---
title: "Outliers"
output: html_document
---

----ANDREA's analysis begins-----

# Outliers Brisbane
```{r}
library(chemometrics)
library(DMwR2)
library(missForest)
library(VIM)
library(ggplot2)
library(GGally)
library(car)

df <- read.csv("Location CSVs/df_Brisbane_MICE.csv")

```

We use only the numeric variables:

```{r}
names(df)
df.active <- df[, c("MinTemp", "MaxTemp", "Rainfall", "Evaporation", "Sunshine", "WindGustSpeed", "WindSpeed9am", "WindSpeed3pm", "Humidity9am", "Humidity3pm", "Pressure9am", "Pressure3pm", "Cloud9am", "Cloud3pm", "Temp9am", "Temp3pm")]
```

# DETECTING UNIVARIATE OUTLIERS BOXPLOT

```{r}

invisible(lapply(1:ncol(df.active), function(i) Boxplot(df.active[, i], ylab='', main=paste("Boxplot", names(df.active)[i], sep=" "), col='orange3', id=list(location='avoid'))))

out <- boxplot.stats(df.active$Pressure9am)$out
out_ind <- which(df.active$Pressure9am %in% c(out))
out_ind
Y<- data.frame(df.active[out_ind, ],df[out_ind,1])
Y

median(df$Rainfall)
```

We can observe that the principal outliers are:
- Boxplot MaxTemp observation 1044
- Boxplot Rainfall observations <b>1274, 1612</b>, 597, 924...
- Boxplot Evaporation observations 472, 146, 833, 895 and 1261
- Boxplot WindGustSpeed observation <b>470</b>, 143, 1611, 1571, 1612, 146, 773...
- Boxplot WindSpeed9am observation <b>1612</b>, 838, 146, 1535, 471...
- Boxplot WindSpeed3pm observations 471, 773, 152, 146
- Boxplot Humidity9am observations 471, 474, 422, 454, 472, 473, 1501, 452...
- Boxplot Humidity3pm observations 422, 450, 453, 1368, 1617, 471, 455, 454...
- Boxplot Pressure9am observations 1612, 145, 1268, 146, 1230...
- Boxplot Pressure3pm observations 145, 146, 1612, 1593, 1286...
- Boxplot Temp3pm observations 1260, 1044

We can see that the observations with most of the variables with outliers are: 1612, 146, 471. We also need to destacate that the observations 1274 and 1612 are strong outliers of the variable Rainfall, the observation 470 that is a strong outlier of the variable WindGustSpeed, and observation 1612 that is a strong outlier of the variable WindSpeed9am. 

# DETECTING MULTIVARIATE OUTLIERS WITH MAHALANOBIS DSTANCE

```{r}
mdi = mahalanobis(df.active, center=apply(df.active,2,mean),cov=var(df.active))
plot(density(mdi))
cutoff <- qchisq(p = 0.99 , ncol(df.active))
## Display observation whose distance greater than cutoff value
Z1 <- data.frame(df.active[mdi>cutoff,],df[mdi>cutoff,1])
Z1

# We select the ones with the greatest Mahalanobis distance
cutoff<-80
Z2 <- data.frame(df.active[mdi>cutoff,],df[mdi>cutoff,1])
Z2
```
We can see that almost all of the outliers detected with the mahalanobis distance where the same that we detected with the boxplot.

# DETECTING MULTIVARIATE OUTLIERS USING Moutlier function

```{r}
library(chemometrics)
dis <- Moutlier(df.active[,c(1:2,4:16)], quantile = 0.99, plot=FALSE)
str(dis)
par(mfrow=c(1,1))
plot(dis$md,dis$rd)
text(dis$md,dis$rd,labels=rownames(df))
abline(h=dis$cutoff, col="red")
abline(v=dis$cutoff, col="red")

cut<-rep(dis$cutoff,nrow(df.active))
cutoff
# As we have a long dataset we use better the robust distance

plot(dis$md)
text(dis$md,labels=rownames(df))
lines(cut,col=2)
dis$cutoff
b<-which(dis$rd>dis$cutoff)

B1 <- data.frame(df.active[b,],df[b,1])
B1

# We select the ones with the greatest Mahalanobis distance
b<-which(dis$rd>10)

B2 <- data.frame(df.active[b,],df[b,1])
B2
```
We can see that the ones with largest Mahalanobis robust distance are observations 899, 470 and 1612.

# DETECTING MULTIVARIATE OUTLIERS BY DENSITY OF LOCAL DISTANCES
.
```{r}
outlier.scores <- lofactor(df.active, k=5)
plot(density(outlier.scores))
# pick top 3 outliers
outliers <- order(outlier.scores, decreasing=T)[1:3]
outliers
sort(outlier.scores,decreasing=T)[1:3]
# who are outliers
print(df[outliers,])

```
We select the top 3 outliers that corresponds with observations 470, 1274 and 1612.
We should remove these outliers and then treat them as missing data.

^^^^ANDREA's analysis above ^^^^^

---- KAT's analysis begins below ----

I start using the complete time series data without any NA's.

```{r}
library(mvoutlier)

#LOAD DATA
df <- read.csv("Completed_Location_CSVs/df_Brisbane_completed.csv", stringsAsFactors = T)

#CHANGE VARIABLE TYPE
df$Date <- as.Date(df$Date)

str(df)

```

To account for weather differences in seasons, I divide the dataset by season for
outlier detection. Also, different groups of variables may be correlated based on
the time of collection, so I create groups of variables for analysis.

Rainfall and TempRange features raise errors when working with the mvoutlier library,
so they are excluded.

```{r}
#split data by season for outlier detection
autumn <- subset(df, Season == "autumn")
spring <- subset(df, Season == "spring")
summer <- subset(df, Season == "summer")
winter <- subset(df, Season == "winter")

#identify numeric variables excluding Rainfall and TempRange
numvar <- c("WindSpeed9am","Humidity9am","Pressure9am","Cloud9am","Temp9am",
            "WindSpeed3pm","Humidity3pm", "Pressure3pm","Cloud3pm","Temp3pm",
            "MinTemp", "MaxTemp", "Evaporation", "Sunshine", "WindGustSpeed")

#identify morning variables
morning <- c("WindSpeed9am","Humidity9am","Pressure9am","Cloud9am","Temp9am")

#identify afternoon variables
afternoon <- c("WindSpeed3pm","Humidity3pm", "Pressure3pm","Cloud3pm","Temp3pm")
  
#identify all day variables excluding TempRange, Rainfall
allDay <- c("MinTemp", "MaxTemp", "Evaporation","Sunshine", "WindGustSpeed")

```

Since detecting outliers has shown to be useful for identifying days with rainfall,
an interesting metric will be to compare the number of rainy days included in
outliers compared with the number of rainy days in the complete data set.

```{r}
#FULL DATA
#number of days with rain today
length(which(df$RainToday == 'Yes'))
#number of days with rain tomorrow (should match rain today)
length(which(df$RainTomorrow == 'Yes'))
#percentage of days with rain today in data set
length(which(df$RainToday == 'Yes'))/nrow(df)*100
#percentage of days with rain tomorrow in data set (should match rain today)
length(which(df$RainTomorrow == 'Yes'))/nrow(df)*100
```

Brisbane RainToday & RainTomorrow numbers do not match which could indicate a problem
with the imputation.


The color.plot function is interesting, but can only be used for 2 dimensions.

Humidity9am and Humidity3pm are correlated and may be interesting to look at.
A change in humidity could be related to rainfall.

```{r}
autHum <- autumn[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(autHum, quan=1/2, alpha=0.025, sub='Autumn Humidity9am vs Humidity3pm')
dfAutHumOutliers <- autumn[which(colorData$outliers),]

sprHum <- spring[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(sprHum, quan=1/2, alpha=0.025, sub='Spring Humidity9am vs Humidity3pm')
dfSprHumOutliers <- spring[which(colorData$outliers),]

sumHum <- summer[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(sumHum, quan=1/2, alpha=0.025, sub='Summer Humidity9am vs Humidity3pm')
dfSumHumOutliers <- summer[which(colorData$outliers),]

winHum <- winter[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(winHum, quan=1/2, alpha=0.025, sub='Winter Humidity9am vs Humidity3pm')
dfWinHumOutliers <- winter[which(colorData$outliers),]

HumidityOutliers <- rbind(dfAutHumOutliers, dfSprHumOutliers, dfSumHumOutliers, dfWinHumOutliers)
```


```{r}
#number of rainy days in the Humidity outliers data
length(which(HumidityOutliers$RainToday == 'Yes'))
#percentage of days with rainfall in the Humidity outliers data
length(which(HumidityOutliers$RainToday == 'Yes'))/nrow(HumidityOutliers)*100

#number of day with rain tomorrow in the Humidity outliers data
length(which(HumidityOutliers$RainTomorrow == 'Yes'))
#percentage of days with rain tomorrow in the Humidity outliers data
length(which(HumidityOutliers$RainTomorrow == 'Yes'))/nrow(HumidityOutliers)*100
```

