---
title: "Outliers"
output: html_document
---
----ANDREA's analysis begins-----

# Outliers Cairns
```{r}
library(chemometrics)
library(DMwR2)
library(missForest)
library(VIM)
library(ggplot2)
library(GGally)
library(car)

df <- read.csv("Location CSVs/df_Cairns_MICE.csv")

```

We use only the numeric variables:

```{r}
names(df)
df.active <- df[, c("MinTemp", "MaxTemp", "Rainfall", "Evaporation", "Sunshine", "WindGustSpeed", "WindSpeed9am", "WindSpeed3pm", "Humidity9am", "Humidity3pm", "Pressure9am", "Pressure3pm", "Cloud9am", "Cloud3pm", "Temp9am", "Temp3pm")]
```

# DETECTING UNIVARIATE OUTLIERS BOXPLOT

```{r}

invisible(lapply(1:ncol(df.active), function(i) Boxplot(df.active[, i], ylab='', main=paste("Boxplot", names(df.active)[i], sep=" "), col='green3', id=list(location='avoid'))))

out <- boxplot.stats(df.active$Pressure9am)$out
out_ind <- which(df.active$Pressure9am %in% c(out))
out_ind
Y<- data.frame(df.active[out_ind, ],df[out_ind,1])
Y

median(df$Rainfall)

df_ordenado <- df.active[order(df.active$WindSpeed9am), ]
head(df_ordenado)
df_ordenado
```

We can observe that the principal outliers are:
- Boxplot MinTemp observations 901, 963, 935, 936, 1321, 898...
- Boxplot MaxTemp observations 1460, 1458, 431, 1242, 910, 177...
- Boxplot Rainfall observations <b>43, 796</b>, 1023, 817, 1176...
- Boxplot Evaporation observations 752, 198, 638, 1235, 1200, 1177...
- Boxplot WindGustSpeed observation <b>794</b>, 1454, 1504, 795, 420, 1534, 1484
- Boxplot WindSpeed9am observation 1524, 794, 286, 689, 615...
- Boxplot WindSpeed3pm observations 469, 987, 944, 821, 455, 415
- Boxplot Humidity9am observations 1320, 309, 984, 1321, 955
- Boxplot Humidity3pm observations 818, 795, 96, 298, 309...
- Boxplot Pressure9am observations 1454, 1130, 1455, 1129, 1131...
- Boxplot Pressure3pm observations 1130, 1454, 1129, 1455, 1456...
- Boxplot Temp9am observations 898, 934, 901
- Boxplot Temp3pm observations 242, 948

We can see that the observations with most of the variables with outliers is 1454. We also need to destacate that the observations 43 and 796 are strong outliers of the variable Rainfall, and the observation 794 that is a strong outlier of the variable WindGustSpeed. 

# DETECTING MULTIVARIATE OUTLIERS WITH MAHALANOBIS DSTANCE

```{r}
mdi = mahalanobis(df.active, center=apply(df.active,2,mean),cov=var(df.active))
plot(density(mdi))
cutoff <- qchisq(p = 0.99 , ncol(df.active))
## Display observation whose distance greater than cutoff value
Z1 <- data.frame(df.active[mdi>cutoff,],df[mdi>cutoff,1])
Z1

# We select the ones with the greatest Mahalanobis distance
cutoff<-65
Z2 <- data.frame(df.active[mdi>cutoff,],df[mdi>cutoff,1])
Z2
```
We can see that almost all of the outliers detected with the mahalanobis distance where the same that we detected with the boxplot.

# DETECTING MULTIVARIATE OUTLIERS USING Moutlier function

```{r}
library(chemometrics)
dis <- Moutlier(df.active[,c(1:2,4:16)], quantile = 0.99, plot=FALSE)
str(dis)
par(mfrow=c(1,1))
plot(dis$md,dis$rd)
text(dis$md,dis$rd,labels=rownames(df))
abline(h=dis$cutoff, col="red")
abline(v=dis$cutoff, col="red")

cut<-rep(dis$cutoff,nrow(df.active))
cutoff
# As we have a long dataset we use better the robust distance

plot(dis$md)
text(dis$md,labels=rownames(df))
lines(cut,col=2)
dis$cutoff
b<-which(dis$rd>dis$cutoff)

B1 <- data.frame(df.active[b,],df[b,1])
B1

# We select the ones with the greatest Mahalanobis distance
b<-which(dis$rd>9.5)

B2 <- data.frame(df.active[b,],df[b,1])
B2
```
We can see that the ones with largest Mahalanobis robust distance is observation 794.

# DETECTING MULTIVARIATE OUTLIERS BY DENSITY OF LOCAL DISTANCES
.
```{r}
outlier.scores <- lofactor(df.active, k=5)
plot(density(outlier.scores))
# pick top 3 outliers
outliers <- order(outlier.scores, decreasing=T)[1:3]
outliers
sort(outlier.scores,decreasing=T)[1:3]
# who are outliers
print(df[outliers,])

```
We select the top 3 outliers that corresponds with observations 794, 43 and 796
We should remove these outliers and then treat them as missing data. 

^^^^ANDREA's analysis above ^^^^^

---- KAT's analysis begins below ----

I start using the complete time series data without any NA's.

```{r}
library(mvoutlier)

#LOAD DATA
df <- read.csv("Completed_Location_CSVs/df_Cairns_completed.csv", stringsAsFactors = T)

#CHANGE VARIABLE TYPE
df$Date <- as.Date(df$Date)

str(df)

```

To account for weather differences in seasons, I divide the dataset by season for
outlier detection. Also, different groups of variables may be correlated based on
the time of collection, so I create groups of variables for analysis.

Rainfall and TempRange features raise errors when working with the mvoutlier library,
so they are excluded.

```{r}
#split data by season for outlier detection
wet <- subset(df, Season == "wet")
dry <- subset(df, Season == "dry")


#identify numeric variables excluding Rainfall and TempRange
numvar <- c("WindSpeed9am","Humidity9am","Pressure9am","Cloud9am","Temp9am",
            "WindSpeed3pm","Humidity3pm", "Pressure3pm","Cloud3pm","Temp3pm",
            "MinTemp", "MaxTemp", "Evaporation", "Sunshine", "WindGustSpeed")

#identify morning variables
morning <- c("WindSpeed9am","Humidity9am","Pressure9am","Cloud9am","Temp9am")

#identify afternoon variables
afternoon <- c("WindSpeed3pm","Humidity3pm", "Pressure3pm","Cloud3pm","Temp3pm")
  
#identify all day variables excluding TempRange, Rainfall
allDay <- c("MinTemp", "MaxTemp", "Evaporation","Sunshine", "WindGustSpeed")

```


Since detecting outliers has shown to be useful for identifying days with rainfall,
an interesting metric will be to compare the number of rainy days included in
outliers compared with the number of rainy days in the complete data set.

```{r}
#FULL DATA
#number of days with rain today
length(which(df$RainToday == 'Yes'))
#number of days with rain tomorrow (should match rain today)
length(which(df$RainTomorrow == 'Yes'))
#percentage of days with rain today in data set
length(which(df$RainToday == 'Yes'))/nrow(df)*100
#percentage of days with rain tomorrow in data set (should match rain today)
length(which(df$RainTomorrow == 'Yes'))/nrow(df)*100

#WET SEASON
#number of days with rain today
length(which(wet$RainToday == 'Yes'))
#number of days with rain tomorrow 
length(which(wet$RainTomorrow == 'Yes'))
#percentage of days with rain today in data set
length(which(wet$RainToday == 'Yes'))/nrow(wet)*100
#percentage of days with rain tomorrow in data set 
length(which(wet$RainTomorrow == 'Yes'))/nrow(wet)*100

#DRY SEASON
#number of days with rain today
length(which(dry$RainToday == 'Yes'))
#number of days with rain tomorrow 
length(which(dry$RainTomorrow == 'Yes'))
#percentage of days with rain today in data set
length(which(dry$RainToday == 'Yes'))/nrow(dry)*100
#percentage of days with rain tomorrow in data set 
length(which(dry$RainTomorrow == 'Yes'))/nrow(dry)*100
```

33% of days are rainy in Cairns overall.
47% of days in the wet season are rainy.
16% of days in the dry season are rainy.



The color.plot function is interesting, but can only be used for 2 dimensions.

Humidity9am and Humidity3pm are correlated and may be interesting to look at.
A change in humidity could be related to rainfall.

```{r}
wetHum <- wet[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(wetHum, quan=1/2, alpha=0.025, sub='Wet Season Humidity9am vs Humidity3pm')
dfWetHumOutliers <- wet[which(colorData$outliers),]

dryHum <- dry[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(dryHum, quan=1/2, alpha=0.025, sub='Dry Season Humidity9am vs Humidity3pm')
dfDryHumOutliers <- dry[which(colorData$outliers),]

HumidityOutliers <- rbind(dfDryHumOutliers, dfWetHumOutliers)

```

```{r}
#number of rainy days in the Humidity outliers data
length(which(HumidityOutliers$RainToday == 'Yes'))
#percentage of days with rainfall in the Humidity outliers data
length(which(HumidityOutliers$RainToday == 'Yes'))/nrow(HumidityOutliers)*100

#number of day with rain tomorrow in the Humidity outliers data
length(which(HumidityOutliers$RainTomorrow == 'Yes'))
#percentage of days with rain tomorrow in the Humidity outliers data
length(which(HumidityOutliers$RainTomorrow == 'Yes'))/nrow(HumidityOutliers)*100
```

Overall, in the set of data recognized as Humidity Outliers, 51% of them have Rain Today
and 57% of them have Rain Tomorrow. That is quite a bit higher than the 33% of rainy days
in the overall Cairns data set.

Next, we can evaluate if there is a difference in Humidity Outliers in the dry
and wet seasons.

```{r}
#number of rainy days in the Humidity outliers dry season data
length(which(dfDryHumOutliers$RainToday == 'Yes'))
#percentage of days with rainfall in the Humidity outliers dry season data
length(which(dfDryHumOutliers$RainToday == 'Yes'))/nrow(dfDryHumOutliers)*100

#number of day with rain tomorrow in the Humidity outliers dry season data
length(which(dfDryHumOutliers$RainTomorrow == 'Yes'))
#percentage of days with rain tomorrow in the Humidity outliers dry season data
length(which(dfDryHumOutliers$RainTomorrow == 'Yes'))/nrow(dfDryHumOutliers)*100
```

In the humidity outliers for the dry season, 55% of them have Rain Tomorrow,
which is substantially higher compared with 16% of overall days in the dry season 
which have Rain Tomorrow.


```{r}
#number of rainy days in the Humidity outliers data
length(which(dfWetHumOutliers$RainToday == 'Yes'))
#percentage of days with rainfall in the Humidity outliers data
length(which(dfWetHumOutliers$RainToday == 'Yes'))/nrow(dfWetHumOutliers)*100

#number of day with rain tomorrow in the Humidity outliers data
length(which(dfWetHumOutliers$RainTomorrow == 'Yes'))
#percentage of days with rain tomorrow in the Humidity outliers data
length(which(dfWetHumOutliers$RainTomorrow == 'Yes'))/nrow(dfWetHumOutliers)*100
```

The wet season has fewer outliers for humidity overall, but 61% of the outliers are rainy, 
which is higher than the 47% of total days in the wet season that are rainy.

