---
pdf_document:
  latex_engine: xelatex
  includes:
    in_header: C:/Users/Diana Rocio/Documents/R/win-library/4.1/summarytools/includes/fig-valign.tex
header-includes:
- \usepackage{multirow}
- \usepackage{longtable}
title: 'Group 1: Multivariate analysis of australian climate data'
output: pdf_document
---


```{r setup, message=FALSE, warning=FALSE, results=FALSE,include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# CLEARING ENVIRONMENT #
rm(list=ls(all=TRUE))


#INSTALL & LOAD LIBRARIES
# Required packages
pkgs<-c("cluster","colorspace","dplyr","factoextra","fpc","ggplot2","ggpubr",
        "NbClust","patchwork","reshape2","stats","summarytools","tidyr","tidyverse","rstudioapi")

# Non-installed packages
inspkgs<-pkgs[!pkgs %in% installed.packages()]
for(libs in inspkgs) install.packages(libs, repos = "http://cran.us.r-project.org")

# Loading required
sapply(pkgs,require,character=TRUE)
```

## Data input

To perform the clustering analysis are used the original datasets (numeric variables) for Brisbane, Perth and Cairns.

```{r pressure, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Loading dataset
load('BrisbaneMFA.RData')
dfBrisbane<-df
dfMFABrisbane<-df_MFA
#EigValMFABrisbane<-eig.val
#groupMFABrisbane<-group
#quali.varMFABrisbane<-quali.var
#quanti.varMFABrisbane<-quanti.var
#resMFABrisbane<-get_mfa_ind(res.mfa)
#coordMFABrisbane<-data.frame(resMFABrisbane$coord)

load('CairnsMFA.RData')
dfCairns<-df
dfMFACairns<-df_MFA
#resMFACairns<-get_mfa_ind(res.mfa)
#coordMFACairns<-data.frame(resMFACairns$coord)

load('PerthMFA.RData')
dfPerth<-df
dfMFAPerth<-df_MFA
#resMFAPerth<-get_mfa_ind(res.mfa)
#coordMFAPerth<-data.frame(resMFAPerth$coord)

originaldata<-list(dfMFABrisbane,dfMFAPerth,dfMFACairns)
funum <- function(x){nums<-unlist(lapply(x, is.numeric))
                     x[ , nums]}
names(originaldata)<-c("Brisbane","Perth","Cairns")


orNUMdata<-lapply(originaldata,funum)
names(orNUMdata)<-c("Brisbane","Perth","Cairns")

Scaled<-lapply(orNUMdata,as.data.frame(scale))
str(Scaled)
names(Scaled)<-c("ScaledBrisbane","ScaledPerth","ScaledCairns")

funames<-function(x){ tmp_df<-x
                      colnames(tmp_df)<-c("WindGustSpeed","WindSpeed9am","WindSpeed3pm",
                                    "MinTemp","MaxTemp","TempRange","Temp9am","Temp3pm",
                                    "Evaporation","Sunshine","Cloud9am","Cloud3pm",
                                    "Rainfall","Humidity9am","Humidity3pm",
                                    "Pressure9am","Pressure3pm")
                    return(tmp_df)}
scaled<-lapply(Scaled, funames)
Scaled<-scaled

listall<-list(orNUMdata$Brisbane,Scaled$ScaledBrisbane,orNUMdata$Perth,Scaled$ScaledPerth,orNUMdata$Cairns,Scaled$ScaledCairns)
names(listall)<-c("Brisbane","ScaledBrisbane","Perth","ScaledPerth","Cairns","ScaledCairns")
str(listall)


set.seed(2021)
```

###  Clustering Analysis

In order to analyze if data presents patterns of association are it is performed a clustering analysis. For this purpose, all incomplete cases remaining are removed and as a first step, the optimal number of clusters are estimated through direct methods: elbow, average silhouette and ASM to choose the most common value of optimal clusters.


```{r, message=FALSE, warnings=FALSE, fig.show="hold", out.width="50%", results=FALSE}
par(mar = c(4,4,.1,.1))
fun01<-function(x){ tmp_df = listall[[x]]
                    tmp_name = names(listall)[x]
                    fviz_nbclust(tmp_df, kmeans, method = "wss") + 
                    geom_vline(xintercept = 4, linetype = 2) + 
                    labs(subtitle = paste("Elbow method",tmp_name))}
fun02<-function(x){ tmp_df = listall[[x]]
                    tmp_name = names(listall)[x]
                    fviz_nbclust(tmp_df, kmeans, method = "silhouette") + 
                    geom_vline(xintercept = 4, linetype = 2) + 
                    labs(subtitle = paste("Elbow method",tmp_name))}
fun03<-function(x){ tmp_df = listall[[x]]
                    tmp_name = names(listall)[x]
                    fviz_nbclust(tmp_df, kmeans, method = "gap_stat") +
                    geom_vline(xintercept = 4, linetype = 2) + 
                    labs(subtitle = paste("Elbow method",tmp_name))}

wss<-lapply(1:length(listall),fun01)
wss

silhouette<-lapply(1:length(listall),fun02)
silhouette

#Gaps<-lapply(1:length(listall),fun03)
#Gaps
```
Given the results provided by the methods, it can be concluded the clustering can be performed with 4 cluster for all the dataset, the original numerical variables and the coordinates of the performed MCA.

```{r, message=FALSE, warnings=FALSE, fig.show="hold", out.width="50%", results=FALSE}
funVizKm<- function(i){ tmp_df = listall[[i]];
                        tmp_kmeans = kmeans(x = listall[[i]], centers = 4)
                        tmp_name = names(listall)[i]
                        fviz_cluster(object = tmp_kmeans, data = listall[[i]], 
                                     show.clust.cent = TRUE, ellipse.type = "euclid",
                                     star.plot = TRUE, repel = TRUE) + 
                     theme_bw() + theme(legend.position = "none") + 
                          labs(title = paste("Results clustering K-means (4 clusters)",
                                             tmp_name))}  

VizKmeans<-lapply(1:length(listall),funVizKm) 
VizKmeans

```
```{r, message=FALSE, warnings=FALSE, fig.show="hold"}
funKm<- function(i){ tmp_df = listall[[i]];
                     tmp_kmeans = kmeans(x = listall[[i]], centers = 4)
                     listall[[i]]<-add_column(listall[[i]], KmeansCluster = 
                                                tmp_kmeans$cluster)}

Kmeans<-lapply(1:length(listall),funKm)
names(Kmeans)<-c("Brisbane","ScaledBrisbane","Perth","ScaledPerth",
                 "Cairns","ScaledCairns")

#a<-Kmeans[[1]]

#fun08<-function(a,i) {
#  for (i in 1:17) {boxplot(a[,i] ~ a[,18], xlab = 'Kmeans', ylab = names(a)[i])}}

#lapply(Kmeans,fun08)

#boxplot(a[,1] ~ a[,18])
#boxplot(a[,2] ~ a[,18])

#boxplot(a)

#plot(formula = KmeansCluster ~ ., data = a)


lapply(1:length(Kmeans), function(x){

  # Get the dataframe and the name 
  tmp_df = Kmeans[[x]]
  tmp_name = names(Kmeans)[x]
  
  for (i in 1:17) {boxplot(tmp_df[,i] ~ tmp_df[,18], xlab = 'Kmeans', ylab = tmp_name, main = tmp_name)}
  })


```

```{r, message=FALSE, warnings=FALSE, fig.show="hold", out.width="50%", results=FALSE}
fun04<-function(x) print(names(x))
lapply(Kmeans, fun04)

fun05<-function(x){x[,ncol(x)]
                   x$KMCluster<-x[,ncol(x)]
                   return(x$KMCluster)}

clusters<-lapply(Kmeans,fun05)

BrisbaneClusters<-as.data.frame(cbind(originaldata[[1]],as.factor(clusters[[1]]),
                                      as.factor(clusters[[2]])))
names(BrisbaneClusters)<-c(names(originaldata[[1]]),"KmeansDF","KmeansScaled")

PerthClusters<-as.data.frame(cbind(originaldata[[2]],as.factor(clusters[[3]]),
                                   as.factor(clusters[[4]])))

names(PerthClusters)<-c(names(originaldata[[3]]),"KmeansDF","KmeansScaled")

CairnsClusters<-as.data.frame(cbind(originaldata[[3]],as.factor(clusters[[5]]),
                                    as.factor(clusters[[6]])))
names(CairnsClusters)<-c(names(originaldata[[3]]),"KmeansDF","KmeansScaled")



funMetrics<-function(i){ tmp_df = listall[[i]]

                     }
lapply(1:length(listall), funMetrics)



```

```{r, message=FALSE, warning=FALSE}
write.csv(BrisbaneClusters,"BrisbaneClusters.csv")
write.csv(PerthClusters,"PerthClusters.csv")
write.csv(CairnsClusters,"CairnsClusters.csv")

DFClusters<-list(BrisbaneClusters,PerthClusters,CairnsClusters)


fun06<-function(x){tmpdf=DFClusters[[x]]
                   levels(tmpdf[,24])<-list(C1="1",C2="2",C3="3",C4="4")
                   levels(tmpdf[,25])<-list(G1="1",G2="2",G3="3",G4="4")
                   return(tmpdf)}
data<-lapply(1:length(DFClusters),fun06)

```



```{r, message=FALSE, warnings=FALSE, fig.show="hold"}
funtableKmeans<-function(x){table(x$KmeansDF,x$KmeansScaled)}
funtabseason<-function(x){table(x$KmeansDF,x$Season) }
funtabseason2<-function(x){table(x$KmeansScaled,x$Season) }
funtabseason2<-function(x){table(x$KmeansScaled,x$RainTomorrow) }

lapply(data, funtableKmeans)
lapply(data,funtabseason)
lapply(data,funtabseason2)

```




```{r, message=FALSE, warnings=FALSE, fig.show="hold"}
#boxplot(BrisbaneClusters$Pressure9am ~ BrisbaneClusters$KmeansDF)

funtableKmeans<-function(x){table(x$KmeansDF,x$KmeansScaled)}
funtabseason<-function(x){table(x$KmeansDF,x$Season) }
funtabseason2<-function(x){table(x$KmeansScaled,x$Season) }
funtabseason2<-function(x){table(x$KmeansScaled,x$Season) }

lapply(data, funtableKmeans)
lapply(data,funtabseason)
lapply(data,funtabseason2)

```



```{r, message=FALSE, warnings=FALSE, fig.show="hold"}

lapply(1:length(Kmeans), function(x){

  # Get the dataframe and the name 
  tmp_df = Kmeans[[x]]
  tmp_name = names(Kmeans)[x]
  
  for (i in 1:17) {
                    kruskal.test(tmp_df[,13] ~ tmp_df[,18], data = tmp_df)
                    pairwise.wilcox.test(tmp_df[,13], tmp_df[,18], p.adjust.method = "BH")
                    #boxplot(tmp_df[,i] ~ tmp_df[,18], xlab = 'Kmeans', ylab = names(a)[i], main = tmp_name)
                  }
  })

#kruskal.test(Murder ~ res.km.cluster, data = results)
# As the p-value is less than the significance level 0.05, we can conclude that there are significant differences between groups in relation to "Murder".

#From the output of the Kruskal-Wallis test, we know that there is a significant difference between groups, but we don’t know which pairs of groups are different.

#pairwise.wilcox.test(results$Murder, results$res.km.cluster, p.adjust.method = "BH")




funProfile<-function(x){catdes(x, num.var=18, prob = 0.01)
                        catdes(x, num.var=19, prob = 0.01)}


#lapply(temp,funProfile)
```