---
title: 'Group 1: Multivariate analysis of australian climate data'
author: Andrea Iglesias Munilla, Kathryn Weissman, Diana Galindo Gonzalez, Mateo Jacome
  Gonzalez y Pedro Gonzalez Prado
output:
  html_document:
    self_contained: false
editor_options: 
  chunk_output_type: console
---

## Data imputation


```{r, ini, message=FALSE, warning=FALSE, results='hide'}

# CLEARING ENVIRONMENT #
#rm(list=ls(all=TRUE))
#PREPARE WORK DIRECTORY 
#current_path <- getActiveDocumentContext()$path 
#setwd(dirname(current_path ))

#INSTALL & LOAD LIBRARIES
# Required packages
pkgs<-c("corrplot","DT", "dplyr","FactoMineR","factoextra", "ggplot2","lmtest", "grid", "gridExtra", "lubridate", "mice", "sm","summarytools", "VIM", "visdat")

# Non-installed packages
inspkgs<-pkgs[!pkgs %in% installed.packages()]
for(libs in inspkgs) install.packages(libs, repos = "http://cran.us.r-project.org")

# Loading required
sapply(pkgs,require,character=TRUE)

df <- read.csv("Location CSVs/df_Perth.csv",sep=",",stringsAsFactors = T)
df <- subset(df, Location == "Perth")
df$Date <- as.Date(df$Date)
```
###  PERTH

## DETECT MISSING DATA

```{r,  message=FALSE, warning=FALSE, results='hide'}
#Missing Data from WIND already imputed in other .RMd file.

#Detecting missing data by row
mis_ind = rowSums(is.na(df))
m1<-which(mis_ind>0)
df_NAs <- df[m1,]
table(mis_ind)

# Detecting Missing Values by Column as count
mis_col <- colSums(is.na(df))
mis_col


# Missing Data Graphs
vis_miss(df)

aggr(df, col=c('grey','#252525'), numbers=TRUE, sortVars=TRUE, labels=names(df), 
     cex.axis=.5, gap=1, ylab=c("Missing data"," "),border=NA)
```

Very few numerical data are missing, Evaporation is not strongly correlated with any other numeric variable. Cloud3pm is negatively correlated with Sunshine and positively correlated with Cloud9am and Humidity3pm. Sunshine is negatively correlated with the Humidity and Cloud measurements.


```{r,  message=FALSE, warning=FALSE, results='hide'}
#CHECK NAs FOR EVAPORATION
#create variable to identify rows that are missing Evaporation
na_Evaporation <- which(is.na(df_NAs$Evaporation))
#create dataframe to check summary statistics of observations missing Evaporation
df_NA_Evap<- df_NAs[na_Evaporation,]
summary(df_NA_Evap)

#Plotting histograms side by side for rainfall in full data set vs on days when
#Evaporation is NA
c1 <- rgb(173,216,230, max=255, alpha=80, names = "lt.blue") #create transparent color
c2 <- rgb(255,192,203, max=255, alpha=80, names = "lt.pink") # create transparent color

rain_full <- df$Rainfall
rain_evap <- df_NA_Evap$Rainfall
b <- min(c(rain_full,rain_evap), na.rm=TRUE) # Set the minimum for the breakpoints
e <- max(c(rain_full,rain_evap), na.rm=TRUE) # Set the maximum for the breakpoints
ax <- pretty(b:e, n=20) # Make a neat vector for the breakpoints
hgRainfall <- hist(rain_full, breaks = ax, plot = FALSE) # Save first histogram data
hgRainfall_Evap <- hist(rain_evap, breaks = ax, plot = FALSE) # Save 2nd histogram data
plot(hgRainfall, col=c1, freq=FALSE, main="Side by Side Histograms for Rainfall",
     xlab="Rainfall per Day (mm)",) # Plot 1st histogram using a transparent color
plot(hgRainfall_Evap, col=c2, add=TRUE,freq=FALSE) # Add 2nd histogram using different color
legend("topright", c("All Days", "Days with Evaporation NA"), fill=c(c1,c2))

#Plotting histograms side by side for Humidity3pm in full data set vs on days when
#Evaporation is NA

humid3_full <- df$Humidity3pm
humid3_evap <- df_NA_Evap$Humidity3pm
b <- min(c(humid3_full,humid3_evap), na.rm=TRUE) # Set the minimum for the breakpoints
e <- max(c(humid3_full,humid3_evap), na.rm=TRUE) # Set the maximum for the breakpoints
ax <- pretty(b:e, n=20) # Make a neat vector for the breakpoints
hgHumid3 <- hist(humid3_full, breaks = ax, plot = FALSE) # Save first histogram data
hgHumid3_Evap <- hist(humid3_evap, breaks = ax, plot = FALSE) # Save 2nd histogram data
plot(hgHumid3_Evap, col=c2, freq=FALSE, main="Side by Side Histograms for Humidity at 3pm",
     xlab="Relative Humidity Percentage at 3pm",) # Plot 1st histogram using a transparent color
plot(hgHumid3, col=c1, add=TRUE,freq=FALSE) # Add 2nd histogram using different color
legend("topleft", c("All Days", "Days with Evaporation NA"), fill=c(c1,c2))

#Plotting histograms side by side for Humidity9am in full data set vs on days when
#Evaporation is NA

humid9_full <- df$Humidity9am
humid9_evap <- df_NA_Evap$Humidity9am
b <- min(c(humid9_full,humid9_evap), na.rm=TRUE) # Set the minimum for the breakpoints
e <- max(c(humid9_full,humid9_evap), na.rm=TRUE) # Set the maximum for the breakpoints
ax <- pretty(b:e, n=20) # Make a neat vector for the breakpoints
hgHumid9 <- hist(humid9_full, breaks = ax, plot = FALSE) # Save first histogram data
hgHumid9_Evap <- hist(humid9_evap, breaks = ax, plot = FALSE) # Save 2nd histogram data
plot(hgHumid9_Evap, col=c2, freq=FALSE, main="Side by Side Histograms for Humidity at 9am",
     xlab="Relative Humidity Percentage at 9am",) # Plot 1st histogram using a transparent color
plot(hgHumid9, col=c1, add=TRUE,freq=FALSE) # Add 2nd histogram using different color
legend("topleft", c("All Days", "Days with Evaporation NA"), fill=c(c1,c2))


###The missing values for Evaporation are clearly not random. They tend to occur on days
###with higher than average rainfall and humidity. In theory, the higher the relative humidity,
###the lower the evaporation should be.
```


```{r,  message=FALSE, warning=FALSE, results='hide'}
#CHECK NAs FOR SUNSHINE
#create variable to identify rows that are missing Sunshine
na_Sunshine <- which(is.na(df_NAs$Sunshine))
#check rows of observations missing Sunshine
df_NAs[na_Sunshine,]

###The missing values for Sunshine do not appear random. All days are in Summer, 
###and on days with rainfall, high cloud cover, or both.

#CHECK NAs FOR WIND GUST SPEED
#create variable to identify rows that are missing WindGustSpeed
na_GustSpeed <- which(is.na(df_NAs$WindGustSpeed))
#check rows of observations missing WindGustSpeed
df_NAs[na_GustSpeed,]

#Plotting histograms side by side for WindSpeed3pm in full data set vs on days when
#WindGustSpeed is NA

Speed3pm_full <- df$WindSpeed3pm
Speed3pm_wgNA <- df_NAs[na_GustSpeed,]$WindSpeed3pm
b <- min(c(Speed3pm_full,Speed3pm_wgNA), na.rm=TRUE) # Set the minimum for the breakpoints
e <- max(c(Speed3pm_full,Speed3pm_wgNA), na.rm=TRUE) # Set the maximum for the breakpoints
ax <- pretty(b:e, n=20) # Make a neat vector for the breakpoints
hgSpeed3pm <- hist(Speed3pm_full, breaks = ax, plot = FALSE) # Save first histogram data
hgSpeed3pm_wgNA <- hist(Speed3pm_wgNA, breaks = ax, plot = FALSE) # Save 2nd histogram data
plot(hgSpeed3pm_wgNA, col=c2, freq=FALSE, main="Side by Side Histograms for Wind Speed at 3pm",
     xlab="Wind Speed at 3pm",) # Plot 1st histogram using a transparent color
plot(hgSpeed3pm, col=c1, add=TRUE,freq=FALSE) # Add 2nd histogram using different color
legend("topleft", c("All Days", "Days with WindGustSpeed NA"), fill=c(c1,c2))
```


WindGustSpeed is potentially missing randomly, however WindGustDir is only missing when WindGustSpeed is missing.


```{r,  message=FALSE, warning=FALSE, results='hide'}
#CHECK NAs FOR CLOUD
#create variable to identify rows that are missing Cloud9am
na_Cloud9 <- which(is.na(df_NAs$Cloud9am))
#check rows of observations missing Cloud9am
df_NAs[na_Cloud9,]

#create variable to identify rows that are missing Cloud3pm
na_Cloud3 <- which(is.na(df_NAs$Cloud3pm))
#check rows of observations missing Cloud3pm
df_NAs[na_Cloud3,]

###Cloud data is potentially missing randomly. The days that have data missing at 9am
###have data available at 3pm and vice versa.

#IMPUTE MISSING VALUES

###'calm' added to wind levels in other .csv

##create variable to identify rows that are missing wind direction at 9am
#na_WindDir9am <- which(is.na(df$WindDir9am))
##check summary of wind speed at 9am for rows that are missing wind direction at 9am
#summary(df[na_WindDir9am,]$WindSpeed9am)
#### All values of WindDir9am can be imputed to 'Calm'
##imputation using row identifier
#df$WindDir9am[na_WindDir9am] <- 'Calm'

##create variable to identify rows that are missing wind direction at 3pm
#na_WindDir3pm <- which(is.na(df$WindDir3pm))
##check summary of wind speed at 3pm for rows that are missing wind direction at 3pm
#summary(df[na_WindDir3pm,]$WindSpeed3pm)
#### All values of WindDir3pm can be imputed to 'Calm'
##imputation using row identifier
#df$WindDir3pm[na_WindDir3pm] <- 'Calm'

# IMPUTATION BY MICE

md.pattern(df)

tempData <- mice(df[,c("Month","MinTemp","MaxTemp","Rainfall","Evaporation",
                       "Sunshine","WindGustDir","WindGustSpeed","WindDir9am",
                       "WindDir3pm","WindSpeed9am","WindSpeed3pm","Humidity9am",
                       "Humidity3pm","Pressure9am","Pressure3pm","Cloud9am",
                       "Cloud3pm","Temp9am","Temp3pm")],m=3,set.seed(2))
tempData$meth
summary(tempData)
completedData <- complete(tempData,1)
xyplot(tempData,Evaporation ~ Humidity9am+Humidity3pm,pch=18,cex=1)
xyplot(tempData,Evaporation ~ Rainfall,pch=18,cex=1)
densityplot(tempData)

summary(completedData)


#Compare density plots of original data with completed data
plot(density(df$Evaporation, na.rm=TRUE), col='red')
lines(density(completedData$Evaporation), col='blue')

plot(density(df$WindGustSpeed, na.rm=TRUE), col='red')
lines(density(completedData$WindGustSpeed), col='blue')

plot(density(df$Sunshine, na.rm=TRUE), col='red')
lines(density(completedData$Sunshine), col='blue')

plot(density(df$Cloud9am, na.rm=TRUE), col='red')
lines(density(completedData$Cloud9am), col='blue')

plot(density(df$Cloud3pm, na.rm=TRUE), col='red')
lines(density(completedData$Cloud3pm), col='blue')

plot(density(df$Rainfall, na.rm=TRUE), col='red')
lines(density(completedData$Rainfall), col='blue')

###Density plots of imputed variables match density plots of original variables
###with NA's removed.


#Replace original data with completed data from MICE
df[,c("Month","MinTemp","MaxTemp","Rainfall","Evaporation",
      "Sunshine","WindGustDir","WindGustSpeed","WindDir9am",
      "WindDir3pm","WindSpeed9am","WindSpeed3pm","Humidity9am",
      "Humidity3pm","Pressure9am","Pressure3pm","Cloud9am",
      "Cloud3pm","Temp9am","Temp3pm")] <- completedData

#IMPUTE RAINTODAY & RAINTOMORROW

#Fill RainToday & RainTomorrow NAs with appropriate values based on Rainfall.
#identify rows which are missing RainToday & Rainfall > 1 and impute "Yes"
na_RainYes <- which(is.na(df$RainToday) & (df$Rainfall >= 1))
df[na_RainYes,]$RainToday <- "Yes"
#identify dates from imputed values
dates_RainYes <- df[na_RainYes,]$Date
#identify dates before imputed values
dates_RainTomYes <- df[na_RainYes,]$Date - 1 
#identify rows for day before imputed Rain and impute "Yes"
na_RainTomYes <- which(df$Date == dates_RainTomYes)
df[na_RainTomYes,]$RainTomorrow <- "Yes"

summary(df)

#All NAs are filled, write new .csv to use for imputing time series.
#write.csv(df,'df_Perth_MICE.csv', row.names = FALSE)

#https://www.earthdatascience.org/courses/earth-analytics/time-series-data/summarize-time-series-by-month-in-r/
# plot rainfall as time series
df %>%
  ggplot(aes(x = Date, y = Rainfall)) +
  geom_point(color = "darkorchid4") +
  labs(title = "Rainfall",
       y = "Daily rainfall (mm)",
       x = "Date") + theme_bw(base_size = 15)
```

Despite having filled all the NA values, the dataset seems to be lacking records and rows for three whole months. 
It appears we are missing April 2011, December 2012, and February 2013. 

```{r}
#Detecting Missing Dates from date range
date_range <- seq(min(df$Date), max(df$Date), by = 1) 
missing_Dates <- data.frame(Date = date_range[!date_range %in% df$Date]) 

```

While this information is missing in the Kaggle dataset, it is partially available from the original source: http://www.bom.gov.au/climate/data-services/station-data.shtml.
While the missing months are present, we can only access data concerning MaxTemp, MinTemp, and Rainfall. This data, although scarce, can be used to impute many of the variables with a certain confidence. 

```{r}
df$TempRange <- df$MaxTemp - df$MinTemp
cor_matrix = cor(numrain_data[,c(4:20)]) #starting from column 4: MinTemp
corrplot(cor_matrix, method = 'number') # colorful number
```

Thanks to MaxTemp, MinTemp, and TempRange, we can find high correlations with a date's Sunshine, and 9AM and 3 PM measures for Humidity, Pressure, Cloud cover, and Temperature. This means that we can potentially impute those variables for the missing months with a certain fidelity, and from this partially imputed set, find the closest neighbours for imputing wind data and evaporation. First, the data must be gathered and adapted to our schema.

```{r}
#Getting MaxTemp
missing_MaxTemp <- read.csv("CSVs_for_gap_imputation/Perth_MaxTemp.csv",sep=",",stringsAsFactors = T)
missing_MaxTemp$Date <- as.Date(paste(missing_MaxTemp$Year, missing_MaxTemp$Month, missing_MaxTemp$Day, sep="-"))
missing_MaxTemp <- subset(missing_MaxTemp, Date %in% missing_Dates$Date, c(Date, Maximum.temperature..Degree.C.))  
names(missing_MaxTemp) <- c("Date","MaxTemp")
summary(missing_MaxTemp)

        #subset(missing_MaxTemp, Date %in%  df[is.na(df$MaxTemp),]$Date) ##NAs for Maxtemp are shared between the Australian Gov's data and Kaggle's

#Getting MinTemp
missing_MinTemp <- read.csv("CSVs_for_gap_imputation/Perth_MinTemp.csv",sep=",",stringsAsFactors = T)
missing_MinTemp$Date <- as.Date(paste(missing_MinTemp$Year, missing_MinTemp$Month, missing_MinTemp$Day, sep="-"))
missing_MinTemp <- subset(missing_MinTemp, Date %in% missing_Dates$Date, c(Date, Minimum.temperature..Degree.C.))  
names(missing_MinTemp) <- c("Date","MinTemp")
summary(missing_MinTemp)

#Getting rainfall
missing_Rainfall <- read.csv("CSVs_for_gap_imputation/Perth_Rainfall.csv",sep=",",stringsAsFactors = T)
missing_Rainfall$Date <- as.Date(paste(missing_Rainfall$Year, missing_Rainfall$Month, missing_Rainfall$Day, sep="-"))
missing_Rainfall <- subset(missing_Rainfall, Date %in% missing_Dates$Date, c(Date, Rainfall.amount..millimetres.))  
names(missing_Rainfall) <- c("Date","Rainfall")
summary(missing_Rainfall)

#merge
df1 <- merge(x = df, y = missing_Dates, by = "Date", all = T)        #Merge all dates to original df
df2 <- merge(x = missing_MaxTemp, y = missing_MinTemp, by = "Date")  #Merge missing maxtemp with mintemp
df3 <- merge(x = df2, y = missing_Rainfall, by = "Date")             #Merge missing max/minTemp with Rainfall
df4 <- merge(x = df1, y = df3, by = "Date", all.x = T)               #Merge missing variables with original df

#The resulting dataframe after the merges has repeated and split columns. To get rid of them:
df4 <- df4 %>% mutate(MaxTemp = coalesce(MaxTemp.x, MaxTemp.y)) %>%  
  mutate(MinTemp = coalesce(MinTemp.x, MinTemp.y)) %>%
  mutate(Rainfall = coalesce(Rainfall.x, Rainfall.y)) %>%
  select(-c(Rainfall.x, Rainfall.y, MinTemp.x, MinTemp.y, MaxTemp.x, MaxTemp.y))
summary(df4)

#Reconstruct Location, Year, Month, Day, TempRange
df4[, "Year"] <- as.factor(format(df4[,"Date"], "%Y"))
df4[, "Month"] <- as.factor(format(df4[,"Date"], "%B"))
df4[, "Day"] <- as.integer(format(df4[,"Date"], "%d"))

df4[,"Location"] <- "Perth"

df4$TempRange <- df4$MaxTemp - df4$MinTemp

summary(df4)

#Reconstruct RainToday and RainTomorrow
na_RainYes4 <- which(is.na(df4$RainToday) & (df4$Rainfall >= 1))
na_RainNo4 <- which(is.na(df4$RainToday) & (df4$Rainfall < 1))
df4[na_RainYes4,]$RainToday <- "Yes"
df4[na_RainNo4,]$RainToday <- "No"

na_RainTom4 <- which(is.na(df4$RainTomorrow))
dates_na_RainTom4 <- df4[na_RainTom4,]$Date
dates_na_RainTom4_tomorrow <- dates_na_RainTom4 +1
df4[na_RainTom4,]$RainTomorrow <- df4[df4$Date %in% as.Date(dates_na_RainTom4_tomorrow),]$RainToday

df4$monthID <- paste(df4$Year, df4$Month, sep = "-")

summary(df4)

```

Once we've filled and recomputed all the available and logically-imputable variables, we can proceed with MICE again.

```{r}
df4_backup <- df4
df4 <- subset(df4, Month %in% c("diciembre","enero","febrero","marzo","abril"))
summary(df4)

md.pattern(df4)
tempData4 <- mice(df4[,c("Month","MinTemp","MaxTemp","TempRange","Rainfall","Evaporation",
                       "Sunshine","WindGustDir","WindGustSpeed","WindDir9am",
                       "WindDir3pm","WindSpeed9am","WindSpeed3pm","Humidity9am",
                       "Humidity3pm","Pressure9am","Pressure3pm","Cloud9am",
                       "Cloud3pm","Temp9am","Temp3pm")],m=3,set.seed(2))
tempData4$meth
summary(tempData4)

completedData4 <- complete(tempData4,1)

xyplot(tempData4,Evaporation ~ Humidity9am+Humidity3pm,pch=18,cex=1)
xyplot(tempData4,Evaporation ~ Rainfall,pch=18,cex=1)
densityplot(tempData4)

summary(completedData4)


#Replace complete time series data with completed data from MICE
df4[,c("Month","MinTemp","MaxTemp","TempRange","Rainfall","Evaporation",
      "Sunshine","WindGustDir","WindGustSpeed","WindDir9am",
      "WindDir3pm","WindSpeed9am","WindSpeed3pm","Humidity9am",
      "Humidity3pm","Pressure9am","Pressure3pm","Cloud9am",
      "Cloud3pm","Temp9am","Temp3pm")] <- completedData4

summary(df4)

df4$WindGustDir<- factor(df4$WindGustDir, levels = c("N", "NNE", "NE", "ENE", "E", "ESE", 
                                                     "SE", "SSE", "S", "SSW", "SW", "WSW", "W", "WNW", "NW", "NNW", "unkn"))
df4$WindDir9am<- factor(df4$WindDir9am, levels = c("N", "NNE", "NE", "ENE", "E", "ESE", 
                                                   "SE", "SSE", "S", "SSW", "SW", "WSW", "W", "WNW", "NW", "NNW", "calm", "unkn"))
df4$WindDir3pm<- factor(df4$WindDir3pm, levels = c("N", "NNE", "NE", "ENE", "E", "ESE", 
                                                   "SE", "SSE", "S", "SSW", "SW", "WSW", "W", "WNW", "NW", "NNW", "calm", "unkn"))

df4[df4$Date %in% missing_Dates$Date,] <- df4 %>% subset(Date %in% missing_Dates$Date) %>% 
  mutate(WindDir3pm = "unkn", WindDir9am = "unkn", WindGustDir = "unkn")
summary(df4)
df5 <- subset(df4_backup, Month %in% c("mayo","junio","julio","agosto","septiembre","octubre","noviembre"))
df5$monthID <- paste(df5$Year, df5$Month, sep = "-")


df6 <- rbind(df4, df5) %>% arrange(Date)
df6$Season <- ifelse(df6$Month %in% c("diciembre","enero","febrero"), "summer",
              ifelse(df6$Month %in% c("marzo","abril","mayo"), "autumn",
              ifelse(df6$Month %in% c("junio","julio","agosto"), "winter",
              ifelse(df6$Month %in% c("septiembre","octubre","noviembre"), "spring","0"))) 
                     ) 
df6$Season <- as.factor(df6$Season)
summary(df6)
```   

Plot time series to visualize imputed data

```{r}
#Sunshine
sunshine <- ggplot() + geom_point(data = df, aes(x = Date, y = Sunshine), color = "orange")
sunshine <- sunshine + geom_point(data = subset(df4, Date %in% missing_Dates$Date), 
                                  aes(x = Date, y = Sunshine), color = "brown2") + labs(title = "Sunshine",
       y = "Sunshine (h)",
       x = "Date") + theme_bw(base_size = 15)
sunshine

#Evaporation
evaporation <- ggplot() + geom_point(data = df, aes(x = Date, y = Evaporation), color = "orange")
evaporation <- evaporation + geom_point(data = subset(df4, Date %in% missing_Dates$Date), 
                                  aes(x = Date, y = Evaporation), color = "brown2") + labs(title = "Evaporation",
       y = "Evaporation (mm)",
       x = "Date") + theme_bw(base_size = 15)
evaporation

#Pressure
pressure9 <- ggplot() + geom_point(data = df, aes(x = Date, y = Pressure9am), color = "orange")
pressure9 <- pressure9 + geom_point(data = subset(df4, Date %in% missing_Dates$Date), 
                                  aes(x = Date, y = Pressure9am), color = "brown2") + labs(title = "Pressure 9 AM",
       y = "Pressure (hPa)",
       x = "Date") + theme_bw(base_size = 15)
pressure9

#Pressure
pressure3 <- ggplot() + geom_point(data = df, aes(x = Date, y = Pressure3pm), color = "orange")
pressure3 <- pressure3 + geom_point(data = subset(df4, Date %in% missing_Dates$Date), 
                                  aes(x = Date, y = Pressure3pm), color = "brown2") + labs(title = "Pressure 3 AM",
       y = "Pressure (hPa)",
       x = "Date") + theme_bw(base_size = 15)
pressure3

#Humidity
humidity9 <- ggplot() + geom_point(data = df, aes(x = Date, y = Humidity9am), color = "orange")
humidity9 <- humidity9 + geom_point(data = subset(df4, Date %in% missing_Dates$Date), 
                                  aes(x = Date, y = Humidity9am), color = "darkorange3") + labs(title = "Humidity 9 AM",
       y = "Humidity (%)",
       x = "Date") + theme_bw(base_size = 15)
humidity9

#Temperature
temp9 <- ggplot() + geom_point(data = df, aes(x = Date, y = Temp9am), color = "orange")
temp9 <- temp9 + geom_point(data = subset(df4, Date %in% missing_Dates$Date), 
                                  aes(x = Date, y = Temp9am), color = "darkorange3") + labs(title = "Temperature 9 AM",
       y = "Temperature (ºC)",
       x = "Date") + theme_bw(base_size = 15)
temp9

#WindSpeed
wind9 <- ggplot() + geom_point(data = df, aes(x = Date, y = WindSpeed9am), color = "orange")
wind9 <- wind9 + geom_point(data = subset(df4, Date %in% missing_Dates$Date), 
                                  aes(x = Date, y = WindSpeed9am), color = "brown2") + labs(title = "Wind Speed 9 AM",
       y = "Wind Speed (km/h)",
       x = "Date") + theme_bw(base_size = 15)
wind9


```


?as.yearmon()
###PCA

```{r}


df_PCA <- df6[,c(7:29,4,30)]
df_PCA$WindDir3pm <- factor(df_PCA$WindDir9am, levels = c("N", "NNE", "NE", "ENE", "E", "ESE", "SE", "SSE", "S", "SSW", "SW", "WSW", "W", "WNW", "NW", "NNW", "calm","unkn"))
summary(df_PCA)

df_names <- as.data.frame(names(df_PCA))
df_names
activePCA <- c(1,2,4,7:16,19:21)
sup_quant <- c(22) 
sup_qual <- c(3, 5, 6, 17, 18, 23, 24, 25)
names(df_PCA[,activePCA])
names(df_PCA[,sup_quant])
names(df_PCA[,sup_qual])

res.pca = PCA(df_PCA, quanti.sup = sup_quant, quali.sup = sup_qual, scale.unit=TRUE, ncp=5, graph=F) 
```

Visualize PCA results

```{r}
res.pca$eig
fviz_eig(res.pca)

res.pca$quanti.sup
fviz_pca_var(res.pca)
res.pca$quali
fviz_pca_ind(res.pca, habillage = 17, addEllipses =TRUE, #rainToday
             ellipse.type = "confidence", palette = "jco", repel = TRUE) 
#RainToday seems to form a well-clustered cloud of points towards the top left

fviz_pca_ind(res.pca, habillage = 18, addEllipses =TRUE, #rainTomorrow
             ellipse.type = "confidence", palette = "jco", repel = TRUE)
#RainTomorrow behaves similarily as RainToday: rainy days are clustered in the
#humid and colder days.

fviz_pca_ind(res.pca, habillage = 6, addEllipses =TRUE, #winddir
             ellipse.type = "confidence", repel = TRUE,  alpha.ind = 0,
             label = "none") +   xlim(-3, 2) + ylim (-2, 3)
#The baricenters for WindDir3pm seem to point out that eastern-component winds
#tend to have on the warmer, dryer side, while western winds tend to be slightly
#colder and definitely more humid. This makes sense given the fact that Perth
#is located in the west coast, so eastern winds come from the dry continental 
#climate, while western winds will bring humidity and rain from the ocean.

fviz_pca_ind(res.pca, addEllipses =F, #Rainfall
             col.ind = df6$Rainfall, repel = TRUE)+
             scale_color_gradient2(low="white", mid="blue", high="red", midpoint=30)
#Rainy days with average precipitations happen mostly in winter, spring and fall
#while days with stronger precipitations have the potential to come in warmer &
#drier days.
    
basic_plot <- fviz_pca_ind(res.pca, label="none")
ggplot(cbind(basic_plot$data,df6[,c("Season","RainTomorrow")]),
aes(x=x,y=y,col=Season,alpha=RainTomorrow, )) + geom_point() + theme_bw() +
  scale_color_manual(values=c("#edae49", "#66a182", "#d1495b", "#00798c")) +
  scale_alpha_manual(values = c(0.2,0.9))
#Seasons are perfectly spread along PC1, with summer days in the right side,
#winter days in the left, and spring and fall are in the middle mixed together.
#This separation perfectly recapitulates the climate type that Perth belongs to,
#often refered to as mediterranean even if located in the antipodes.
#Besides, PC2 seems to clearly separate humid, rainy days (in the top half, with
#a clear preference towards the left of the plane) from dry ones (in the bottom
#half). 

```

