
# MCA


```{r}

library(FactoMineR)
library(factoextra)

df_Cairns <- read.csv("Completed_Location_CSVs/df_Cairns_completed.csv")
df_Brisbane <- read.csv("Completed_Location_CSVs/df_Brisbane_completed.csv")
df_Perth <- read.csv("Completed_Location_CSVs/df_Perth_completed.csv")

```

Subsetting features and applying MCA Analysis
```{r}

df_Cairns.active <- df_Cairns[, c("Month", "WindGustDir", "WindDir9am", "WindDir3pm", "RainToday", "RainTomorrow")]
df_Brisbane.active <- df_Brisbane[, c("Month", "WindGustDir", "WindDir9am", "WindDir3pm", "RainToday", "RainTomorrow")]
df_Perth.active <- df_Perth[, c("Month", "WindGustDir", "WindDir9am", "WindDir3pm", "RainToday", "RainTomorrow")]

res.mca_Cairns <- MCA(df_Cairns.active, graph = FALSE)
res.mca_Brisbane <- MCA(df_Brisbane.active, graph = FALSE)
res.mca_Perth <- MCA(df_Perth.active, graph = FALSE)

```


# Results of MCA function (Visualization and interpretation)
Eingenvalues and Variance:
```{r}

eig.val_Cairns <- get_eigenvalue(res.mca_Cairns)
head(eig.val_Cairns, n=10)
fviz_screeplot(res.mca_Cairns, addlabels = TRUE, ylim = c(0, 10))

eig.val_Brisbane <- get_eigenvalue(res.mca_Brisbane)
head(eig.val_Brisbane, n=10)
fviz_screeplot(res.mca_Brisbane, addlabels = TRUE, ylim = c(0, 10))

eig.val_Perth <- get_eigenvalue(res.mca_Perth)
head(eig.val_Perth, n=10)
fviz_screeplot(res.mca_Perth, addlabels = TRUE, ylim = c(0, 10))

```

The dimensions of the MCA results does not represent the variability of the model adequately. In order to make a good analysis, we should take the first dimensions that add up to at least an 80% of explained variance. In the graphics above we can see that with the first 10 dimensions we get less than the 30% of the variability.

That is why we will not use this type of analysis and we will resort to techniques that use numerical variables such as PCA and MFA that perform a much more convenient analysis with better results.
