---
title: "PerthMFA"
title: 'Group 1: Multivariate analysis of australian climate data'
author: Andrea Iglesias Munilla, Kathryn Weissman, Diana Galindo Gonzalez, Mateo Jacome
  Gonzalez y Pedro Gonzalez Prado
output:
  html_document:
    self_contained: false
editor_options: 
  chunk_output_type: console
---
```{r}

# CLEARING ENVIRONMENT #

rm(list=ls(all=TRUE))

#PREPARE WORK DIRECTORY 

current_path <- getActiveDocumentContext()$path 
setwd(dirname(current_path ))
saveimages <- "PlotsMFA/Perth/"

```

PACKAGES
```{r}

```

```{r}
library(FactoMineR)
library(factoextra)
```

LOAD AND SUMMARIZE DATA


```{r} 

df <- read.csv("Completed_Location_CSVs/df_Perth_completed.csv",sep=",",stringsAsFactors = T)
df$Date <- as.Date(df$Date)
colnames(df)
summary(df)

```


```{r} 
#MFA

df_MFA <- subset(df, ,c("WindGustSpeed", "WindSpeed9am", "WindSpeed3pm", "MinTemp", "MaxTemp", "TempRange", "Temp9am", "Temp3pm", "Evaporation", "Sunshine", "Cloud9am", "Cloud3pm", "Rainfall", "Humidity9am", "Humidity3pm", "Pressure9am", "Pressure3pm",  "WindGustDir","WindDir9am","WindDir3pm", "RainToday","RainTomorrow", "Season"))

res.mfa <- MFA(df_MFA, group = c(3, 5, 4, 5, 3, 2, 1), type = c("s", "s", "s", "s","n", "n", "n"), name.group = c("WindSpeed","Temperature","Sunlight","ExternalFactors", "WindDirection", "RainBoolean", "Season"), num.group.sup = 7, graph = FALSE) #ALWAYS DECLARE "SEASON" AS SUPPLEMENTARY, IT IS ONLY FOR PLOTTING PURPOSES

# Eigenvalues, cummulative variance and scree plot

eig.val <- get_eigenvalue(res.mfa)
head(eig.val)
#png(paste(saveimages,"screeplot1.png",sep=""),width = 600, height = 350)
fviz_screeplot(res.mfa,addlabels = TRUE, title = "Scree plot Perth")
#dev.off()

```

```{r}

res.mfa <- MFA(df_MFA, group = c(3, 5, 4, 5, 3, 2, 1), type = c("s", "s", "s", "s","n", "n", "n"), name.group = c("WindSpeed","Temperature","Sunlight","ExternalFactors", "WindDirection", "RainBoolean", "Season"), num.group.sup = c(5, 7), graph = FALSE) #ALWAYS DECLARE "SEASON" AS SUPPLEMENTARY

# Eigenvalues, cummulative variance and scree plot

eig.val <- get_eigenvalue(res.mfa)
head(eig.val)
#png(paste(saveimages,"screeplot2.png",sep=""),width = 600, height = 350)
fviz_screeplot(res.mfa,addlabels = TRUE, title = "Scree plot Perth")
#dev.off()


```


Graphs of Variables

```{r}

group <- get_mfa_var(res.mfa, "group")
group
#png(paste(saveimages,"VariableGroups.png",sep=""),width = 600, height = 350)
fviz_mfa_var(res.mfa, "group", title ="Variable groups MFA - Perth")
#dev.off()
# Coordinates of groups
group$coord
# Cos2: quality of representation on the factore map
group$cos2

```

Contributions per group to dimensions. 

```{r}

# Contribution to the first dimension
#png(paste(saveimages,"ContribGroupsDim1.png",sep=""),width = 600, height = 350)
fviz_contrib(res.mfa, "group", axes = 1, title = "Contribution of groups to Dim1 - Perth")
#dev.off()
# Contribution to the second dimension
#png(paste(saveimages,"ContribGroupsDim2.png",sep=""),width = 600, height = 350)
fviz_contrib(res.mfa, "group", axes = 2, title = "Contribution of groups to Dim2 - Perth")
#dev.off()
# Contributions to the  dimensions
group$contrib

```

Quantitative variables

```{r}
quanti.var <- get_mfa_var(res.mfa, "quanti.var")
quanti.var

#Coordinates
quanti.var$coord
# Cos2: quality on the factore map
quanti.var$cos2
# Contributions to the dimensions
quanti.var$contrib

```

Correlations between variables and dimensions

```{r}
#png(paste(saveimages,"QuantitativeVariables.png",sep=""),width = 600, height = 350)
fviz_mfa_var(res.mfa, "quanti.var", palette = "jco", 
             col.var.sup = "violet", repel = TRUE, title = "Quantitative Variables MFA - Perth")
#dev.off()
```

Contributions to dimension 1

```{r}
#png(paste(saveimages,"ContribQuantDim1.png",sep=""),width = 600, height = 350)
fviz_contrib(res.mfa, choice = "quanti.var", axes = 1, top = 20,
             palette = "jco", title = "Contribution of quantitative variables to Dim1 - Perth")
#dev.off()

```

Contributions to dimension 2

```{r}
#png(paste(saveimages,"ContribQuantDim2.png",sep=""),width = 600, height = 350)
fviz_contrib(res.mfa, choice = "quanti.var", axes = 2, top = 20,
             palette = "jco", title = "Contribution of quantitative variables to Dim2 - Perth")
#dev.off()
```

Overall contribution

```{r}
#png(paste(saveimages,"ContribQuant.png",sep=""),width = 600, height = 350)
fviz_mfa_var(res.mfa, "quanti.var", col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             col.var.sup = "violet", repel = TRUE,
             geom = c("point", "text"), title = "Contribution Quantitative Variables MFA - Perth")
#dev.off()

```

Quality of representation(cos2)

```{r}
#png(paste(saveimages,"Cos2QuantVariables.png",sep=""),width = 600, height = 350)
fviz_mfa_var(res.mfa, col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             col.var.sup = "violet", repel = TRUE, title = "Cos2 Quantitative Variables MFA - Perth")
#dev.off()

```

Qualitative variables ## RUN THE MFA WITH ALL VARIABLES BEFORE 

```{r}
quali.var <- get_mfa_var(res.mfa, "quali.var")
quali.var

#png(paste(saveimages,"QualiVarCat.png",sep=""),width = 600, height = 350)
fviz_mfa_var(res.mfa, "quali.var", col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             col.var.sup = "violet", repel = TRUE,
             geom = c("point", "text"), title = "Qualitative variables categories MFA - Perth")
#dev.off()
```

Analysis of individuals

```{r}
ind <- get_mfa_ind(res.mfa)
ind
#png(paste(saveimages,"IndsRainTomorrowBool.png",sep=""),width = 600, height = 350)
fviz_mfa_ind(res.mfa, 
             habillage = "RainTomorrow", # color by groups 
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             addEllipses = TRUE, ellipse.type = "confidence", 
             repel = TRUE, invisible = "quali.var", geom = "point", title = "Individuals MFA - Perth")
#dev.off()

#####
#png(paste(saveimages,"IndsRainComparison.png",sep=""),width = 600, height = 350)
fviz_ellipses(res.mfa, c("RainTomorrow", "RainToday"), repel = TRUE, geom = "point", title = "MFA Factor map - Perth")
#dev.off()


```

Labeling observations by seasons

```{r}
#png(paste(saveimages,"IndsSeason.png",sep=""),width = 600, height = 350)
fviz_mfa_ind(res.mfa, 
             habillage = "Season",
             addEllipses = TRUE, ellipse.type = "confidence", 
             repel = TRUE, 
             invisible = "quali.var", geom = "point", title = "Individuals by Season MFA - Perth")
#dev.off()
```

```{r}
#png(paste(saveimages,"IndsSeasonRain.png",sep=""),width = 600, height = 350)
basic_plot <- fviz_mfa_ind(res.mfa, label="none")
ggplot(cbind(basic_plot$data,df[,c("Season","RainTomorrow")]),
aes(x=x,y=y,col=Season,alpha=RainTomorrow, )) + ggtitle("Individuals MFA - Perth") + geom_point() + theme_bw() +
  scale_color_manual(values=c("#edae49", "#66a182", "#d1495b", "#00798c")) +
  scale_alpha_manual(values = c(0.2,0.9))
#dev.off()
```

