---
title: "Perth_Outlier_Detection"
author: "Kat Weissman"
date: "11/6/2021"
output: html_document
---

```{r setup, include=FALSE}
library(mvoutlier)
```


```{r}
#LOAD DATA
df <- read.csv("Completed_Location_CSVs/df_Perth_completed.csv", stringsAsFactors = T)

#CHANGE VARIABLE TYPE
df$Date <- as.Date(df$Date)

#LIST VARIABLES
str(df)
```

## Summary Statistics

```{r}
#SUMMARIZE DATA BY COLUMN
summary(df)
```
```{r}
#split data by season for outlier detection
autumn <- subset(df, Season == "autumn")
spring <- subset(df, Season == "spring")
summer <- subset(df, Season == "summer")
winter <- subset(df, Season == "winter")

#identify numeric variables excluding Rainfall and TempRange
numvar <- c("WindSpeed9am","Humidity9am","Pressure9am","Cloud9am","Temp9am",
            "WindSpeed3pm","Humidity3pm", "Pressure3pm","Cloud3pm","Temp3pm",
            "MinTemp", "MaxTemp", "Evaporation", "Sunshine", "WindGustSpeed")

#identify morning variables
morning <- c("WindSpeed9am","Humidity9am","Pressure9am","Cloud9am","Temp9am")

#identify afternoon variables
afternoon <- c("WindSpeed3pm","Humidity3pm", "Pressure3pm","Cloud3pm","Temp3pm")
  
#identify all day variables excluding TempRange, Rainfall
allDay <- c("MinTemp", "MaxTemp", "Evaporation","Sunshine", "WindGustSpeed")

```

An error occurs when TempRange and Rainfall variables are included in the multivariate outlier detection.

```{r}
#choose features for multivariate outlier detection
autumnNum <- autumn[ , numvar]
autumn9am <- autumn[ , morning]
autumn3pm <- autumn[ , afternoon]
autumnDay <- autumn[ , allDay]

#compare plots when all variables are included vs. a portion of the variables
autNumDist <- dd.plot(autumnNum, quan=1/2, alpha=0.025, sub='Autumn numeric excluding Rainfall & TempRange')
aut9amDist <- dd.plot(autumn9am, quan=1/2, alpha=0.025, sub='Autumn 9am')
aut3pmDist <- dd.plot(autumn3pm, quan=1/2, alpha=0.025, sub='Autumn 3pm')
autDayDist <- dd.plot(autumnDay, quan=1/2, alpha=0.025, sub='Autumn All Day')
```
```{r}
#choose features for multivariate outlier detection
springNum <- spring[ , numvar]
spring9am <- spring[ , morning]
spring3pm <- spring[ , afternoon]
springDay <- spring[ , allDay]

springNumDist <- dd.plot(springNum, quan=1/2, alpha=0.025, sub='Spring numeric excluding Rainfall & TempRange')
spr9amDist <- dd.plot(spring9am, quan=1/2, alpha=0.025, sub='Spring 9am')
spr3pmDist <- dd.plot(spring3pm, quan=1/2, alpha=0.025, sub='Spring 3pm')
sprDayDist <- dd.plot(springDay, quan=1/2, alpha=0.025, sub='Spring All Day')

outliers <- aq.plot(springNum, delta=qchisq(0.975, df = ncol(springNum)),
                    quan = 1/2, alpha = 0.05)
```
```{r}
#choose features for multivariate outlier detection
summerNum <- summer[ , numvar]
summer9am <- summer[ , morning]
summer3pm <- summer[ , afternoon]
summerDay <- summer[ , allDay]

sumNumDist <- dd.plot(summerNum, quan=1/2, alpha=0.025, sub='Summer numeric excluding Rainfall & TempRange')
sum9amDist <- dd.plot(summer9am, quan=1/2, alpha=0.025, sub='Summer 9am')
sum3pmDist <- dd.plot(summer3pm, quan=1/2, alpha=0.025, sub='Summer 3pm')
sumDayDist <- dd.plot(summerDay, quan=1/2, alpha=0.025, sub='Autumn All Day')
```

```{r}
#choose features for multivariate outlier detection
winterNum <- winter[ , numvar]
winter9am <- winter[ , morning]
winter3pm <- winter[ , afternoon]
winterDay <- winter[ , allDay]

wintNumDist <- dd.plot(winterNum, quan=1/2, alpha=0.025, sub='Winter numeric excluding Rainfall & TempRange')
wint9amDist <- dd.plot(winter9am, quan=1/2, alpha=0.025, sub='Winter 9am')
wintpmDist <- dd.plot(winter3pm, quan=1/2, alpha=0.025, sub='Winter 3pm')
wintDayDist <- dd.plot(winterDay, quan=1/2, alpha=0.025, sub='Winter All Day')
```


The follow plots are interesting, but can only be used for 2 dimensions.
MaxTemp and Evaporation are correlated so may be interesting to look at.

```{r}
autTemp <- autumn[,c("MaxTemp","Evaporation")]
colorData <- color.plot(autTemp, quan=1/2, alpha=0.025, sub='Summer MaxTemp vs Evaporation')
autTempOutliers <- (colorData$outliers)

sprTemp <- spring[,c("MaxTemp","Evaporation")]
colorData <- color.plot(sprTemp, quan=1/2, alpha=0.025, sub='Autumn MaxTemp vs Evaporation')
sprTempOutliers <- (colorData$outliers)

sumTemp <- summer[,c("MaxTemp","Evaporation")]
colorData <- color.plot(sumTemp, quan=1/2, alpha=0.025, sub='Summer MaxTemp vs Evaporation')
sumTempOutliers <- (colorData$outliers)

winTemp <- winter[,c("MaxTemp","Evaporation")]
colorData <- color.plot(winTemp, quan=1/2, alpha=0.025, sub='Winter MaxTemp vs Evaporation')
winTempOutliers <- (colorData$outliers)

```

Humidity9am and Humidity3pm are also correlated so may be interesting to look at.
A change in humidity could be related to rainfall.

```{r}
autHum <- autumn[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(autHum, quan=1/2, alpha=0.025, sub='Autumn Humidity9am vs Humidity3pm')
dfAutHumOutliers <- autumn[which(colorData$outliers),]

sprHum <- spring[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(sprHum, quan=1/2, alpha=0.025, sub='Spring Humidity9am vs Humidity3pm')
dfSprHumOutliers <- spring[which(colorData$outliers),]

sumHum <- summer[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(sumHum, quan=1/2, alpha=0.025, sub='Summer Humidity9am vs Humidity3pm')
dfSumHumOutliers <- summer[which(colorData$outliers),]

winHum <- winter[,c("Humidity9am","Humidity3pm")]
colorData <- color.plot(winHum, quan=1/2, alpha=0.025, sub='Winter Humidity9am vs Humidity3pm')
dfWinHumOutliers <- winter[which(colorData$outliers),]

HumidityOutliers <- rbind(dfAutHumOutliers, dfSprHumOutliers, dfSumHumOutliers, dfWinHumOutliers)

```
Let's check ratio of rainy days by season in the Perth data set

```{r}
ratioRainyDaysAutumn <- length(which(autumn$RainToday == 'Yes'))/nrow(autumn)
numRainyDaysAutumn <- length(which(autumn$RainToday == 'Yes'))

ratioRainyDaysSpring <- length(which(spring$RainToday == 'Yes'))/nrow(spring)
numRainyDaysSpring <- length(which(spring$RainToday == 'Yes'))

ratioRainyDaysSummer <- length(which(summer$RainToday == 'Yes'))/nrow(summer)
numRainyDaysSummer <- length(which(summer$RainToday == 'Yes'))

ratioRainyDaysWinter <- length(which(winter$RainToday == 'Yes'))/nrow(winter)
numRainyDaysWinter <- length(which(winter$RainToday == 'Yes'))
```

Let's check the numbers and percentages of RainTomorrow in the Humidity Outliers
data set versus the total data set for Perth.

```{r}
length(which(HumidityOutliers$RainTomorrow == 'Yes'))
length(which(HumidityOutliers$RainTomorrow == 'Yes'))/nrow(HumidityOutliers)*100

length(which(df$RainTomorrow == 'Yes'))
length(which(df$RainTomorrow == 'Yes'))/nrow(df)*100
```
The number of days where it "rains tomorrow" is significantly higher on the days
with Humidity outliers. 61% of the days in HumidityOutliers have rain tomorrow,
and only 19% of the overall days have rain tomorrow.

Let's check the numbers and ratio of RainToday in the Humidity Outliers
data set versus the total data set for Perth.

```{r}
length(which(HumidityOutliers$RainToday == 'Yes'))
length(which(HumidityOutliers$RainToday == 'Yes'))/nrow(HumidityOutliers)

length(which(df$RainToday == 'Yes'))
length(which(df$RainToday == 'Yes'))/nrow(df)
```

The number of days where it "rains today" is a little bit higher in the HumidityOutliers
but not as much as for rain the following day. 26% of the days in the HumidityOutliers
are rainy versus 19% of days in the overall data set for Perth.

