---
title: "Perth_missing_data"
author: "Kat Weissman"
date: "10/26/2021"
output: html_document
---

```{r setup, include=FALSE}
library (lubridate)
library (dplyr)
library (ggplot2)
library (VIM)
library (visdat)
library (corrplot)
library (grid)
library (gridExtra)
library (sm)
library (mice)
library (tidyverse)
```


```{r}
#LOAD DATA
df <- read.csv("Location CSVs/df_Perth.csv", stringsAsFactors = T)

#CHANGE VARIABLE TYPE
df$Date <- as.Date(df$Date)

#LIST VARIABLES
str(df)
```
```{r}
#FILTER DATA BY LOCATION
df_Airport <- df[which(df$Location == "PerthAirport"),]
df_Perth <- df[which(df$Location == "Perth"),]

#COMPARE DATE RANGES

#Detecting Missing Dates from date range for Perth Airport
date_range_PerthAir <- seq(min(df_Airport$Date), max(df_Airport$Date), by = 1) 
missing_dates_PerthAir <- date_range_PerthAir[!date_range_PerthAir %in% df_Airport$Date] 

#Detecting Missing Dates from date range for Perth City
date_range_Perth <- seq(min(df_Perth$Date), max(df_Perth$Date), by = 1) 
missing_dates_Perth <- date_range_Perth[!date_range_Perth %in% df_Perth$Date]

#Check if dates missing from Perth & PerthAirport are identical
all(missing_dates_PerthAir == missing_dates_Perth)
```

Both locations are missing April 2011, December 2012, and February 2013.
The city of Perth has data available starting 6 months earlier than Perth Airport.
Perth has data starting July 1, 2008 and Perth Airport has data starting on January 1, 2009.

Let's check which location is missing more data.

```{r}
#Check which data is missing from the airport

#Detecting missing data by row
mis_ind_A = rowSums(is.na(df_Airport))
m1_A<-which(mis_ind_A>0)
df_NAs_A <- df_Airport[m1_A,]
table(mis_ind_A)

# Detecting Missing Values by Column as count
mis_col_A <- colSums(is.na(df_Airport))
mis_col_A
```
```{r}
#Check which data is missing from the city of Perth

#Detecting missing data by row
mis_ind_P = rowSums(is.na(df_Perth))
m1_P<-which(mis_ind_P>0)
df_NAs_P <- df_Perth[m1_P,]
table(mis_ind_P)

# Detecting Missing Values by Column as count
mis_col_P <- colSums(is.na(df_Perth))
mis_col_P
```
The city of Perth has more complete data, so going forward, we will use only the Perth location and check use the Airport location as a secondary data set to impute missing values.

```{r}
#Changing variable names and removing duplicates for ease of re-using code.
df_Both <- df
df <- df_Perth
df_NAs <- df_NAs_P
rm (df_Perth)
rm (df_NAs_P)
```

```{r}
#SUMMARIZE DATA BY COLUMN
summary(df)
```

```{r}
#HISTOGRAMS AND CORRELATION PLOTS FOR COMPLETE NUMERIC OBSERVATIONS
nums <- unlist(lapply(df, is.numeric)) 
numrain_data<-df[ , nums]
numrain_data<-subset(numrain_data,complete.cases(numrain_data))
a<-names(numrain_data)
a<-as.list(a)

fun02<-function(i){index=grep(i,names(numrain_data))
bw <- nclass.Sturges(numrain_data[,index]) # Freedman-Diaconis
nm=paste0(i)
assign(paste("g",i,sep=""),
       ggplot(numrain_data, aes(numrain_data[,index])) +  
         geom_histogram(bins = bw,aes(y=..density..), fill="#de2d26") +
         geom_density(alpha=.35, fill="#08519c",color = NA)  +
         geom_vline (aes(xintercept=median(numrain_data[,index])),color="#08519c", size=1) + 
         labs(title=nm, x=NULL, y="UPAS")) +
  theme(plot.title = element_text(size = rel(0.7),face ="bold",hjust = 0.5),
        axis.title.y = element_text(size = rel(0.4)),
        axis.text = element_text(size = rel(0.4)))
}
Histos<-lapply(a[-1],fun02)
do.call(grid.arrange, Histos)

cor_matrix = cor(numrain_data[,c(4:19)]) #starting from column 4: MinTemp
corrplot(cor_matrix, method = 'number', number.cex=0.5) # colorful number

```

# Detecting Missing Data

Missing data from the wind direction was imputed in a previous .r file.

```{r}
#DETECT MISSING DATA
# Missing Data Graphs
vis_miss(df)

aggr(df, col=c('grey','#252525'), numbers=TRUE, sortVars=TRUE, labels=names(df), 
     cex.axis=.5, gap=1, ylab=c("Missing data"," "),border=NA)
```

Humidity 9am is the variable missing the most observations, with 9 missing,
followed by Humidity 3pm with 8 missing. The rest of the variables are fairly complete,
with 4 or fewer missing observations.

```{r}
#DEFINING COLORS FOR SIDE BY SIDE HISTOGRAMS
c1 <- rgb(173,216,230, max=255, alpha=80, names = "lt.blue") #create transparent color
c2 <- rgb(255,192,203, max=255, alpha=80, names = "lt.pink") # create transparent color
```

```{r}
#CHECK NAs FOR HUMIDITY9AM
#create variable to identify rows that are missing Humidity9am
na_Humid9 <- which(is.na(df_NAs$Humidity9am))
#create dataframe to check summary statistics of observations missing Humidity9am
df_NA_Humid9<- df_NAs[na_Humid9,]
summary(df_NA_Humid9)
df_NA_Humid9

#Plotting histograms side by side for Pressure9am in full data set vs on days when
#Humidity9am is NA

Pres9_full <- df$Pressure9am
Pres9_Humid9NA <- df_NA_Humid9$Pressure9am
b <- min(c(Pres9_full,Pres9_Humid9NA), na.rm=TRUE) # Set the minimum for the breakpoints
e <- max(c(Pres9_full,Pres9_Humid9NA), na.rm=TRUE) # Set the maximum for the breakpoints
ax <- pretty(b:e, n=20) # Make a neat vector for the breakpoints
hgPres9 <- hist(Pres9_full, breaks = ax, plot = FALSE) # Save first histogram data
hgPres9_Humid9NA <- hist(Pres9_Humid9NA, breaks = ax, plot = FALSE) # Save 2nd histogram data
plot(hgPres9_Humid9NA, col=c2, freq=FALSE, main="Side by Side Histograms for Pressure9am",
     xlab="Pressure at 9am",) # Plot 1st histogram using a transparent color
plot(hgPres9, col=c1, add=TRUE,freq=FALSE) # Add 2nd histogram using different color
legend("topright", c("All Days", "Days with Humidity9am NA"), fill=c(c1,c2))

```
The Humidity at 9am measure is not missing randomly. There are 5 consecutive days from
December 29, 2008 - January 2, 2009 that are missing the measurement. 4 of those days
are also missing the measurement for Humidity at 3pm.

```{r}
#CHECK NAs FOR HUMIDITY3PM
#create variable to identify rows that are missing Humidity 3pm
na_Humid3 <- which(is.na(df_NAs$Humidity3pm))
#create dataframe to check summary statistics of observations missing Humidity 3pm
df_NA_Humid3<- df_NAs[na_Humid3,]
summary(df_NA_Humid3)
df_NA_Humid3

#Plotting histograms side by side for Pressure9am in full data set vs on days when
#Humidity3pm is NA

Pres9_full <- df$Pressure9am
Pres9_Humid3NA <- df_NA_Humid3$Pressure9am
b <- min(c(Pres9_full,Pres9_Humid3NA), na.rm=TRUE) # Set the minimum for the breakpoints
e <- max(c(Pres9_full,Pres9_Humid3NA), na.rm=TRUE) # Set the maximum for the breakpoints
ax <- pretty(b:e, n=20) # Make a neat vector for the breakpoints
hgPres9 <- hist(Pres9_full, breaks = ax, plot = FALSE) # Save first histogram data
hgPres9_Humid3NA <- hist(Pres9_Humid3NA, breaks = ax, plot = FALSE) # Save 2nd histogram data
plot(hgPres9_Humid3NA, col=c2, freq=FALSE, main="Side by Side Histograms for Pressure9am",
     xlab="Pressure at 9am",) # Plot 1st histogram using a transparent color
plot(hgPres3, col=c1, add=TRUE,freq=FALSE) # Add 2nd histogram using different color
legend("topright", c("All Days", "Days with Humidity3pm NA"), fill=c(c1,c2))

```
Humdity at 3pm is missing 4 days in common with Humidity at 9am, and the other days appear to be
missing randomly. One of the observations on November 9, 2011 is also missing Pressure at 3pm, 
Temperature at 3pm, and Wind Gust Speed.

```{r}
#CHECK NAs FOR SUNSHINE
#create variable to identify rows that are missing Sunshine
na_Sunshine <- which(is.na(df_NAs$Sunshine))
#check rows of observations missing Sunshine
df_NAs[na_Sunshine,]
```
There are two days in a row missing the Sunshine measure. November 27 & 28, 2011. The other days appear
to be missing randomly.

```{r}
#CHECK NAs FOR WIND GUST SPEED
#create variable to identify rows that are missing WindGustSpeed
na_GustSpeed <- which(is.na(df_NAs$WindGustSpeed))
#check rows of observations missing WindGustSpeed
df_NAs[na_GustSpeed,]

#Plotting histograms side by side for WindSpeed3pm in full data set vs on days when
#WindGustSpeed is NA

Speed3pm_full <- df$WindSpeed3pm
Speed3pm_wgNA <- df_NAs[na_GustSpeed,]$WindSpeed3pm
b <- min(c(Speed3pm_full,Speed3pm_wgNA), na.rm=TRUE) # Set the minimum for the breakpoints
e <- max(c(Speed3pm_full,Speed3pm_wgNA), na.rm=TRUE) # Set the maximum for the breakpoints
ax <- pretty(b:e, n=10) # Make a neat vector for the breakpoints
hgSpeed3pm <- hist(Speed3pm_full, breaks=ax, plot = FALSE) # Save first histogram data
hgSpeed3pm_wgNA <- hist(Speed3pm_wgNA,breaks = ax, plot = FALSE) # Save 2nd histogram data
plot(hgSpeed3pm, col=c1, freq=FALSE, main="Side by Side Histograms for Wind Speed at 3pm",
     xlab="Wind Speed at 3pm",) # Plot 1st histogram using a transparent color
plot(hgSpeed3pm_wgNA, col=c2, add=TRUE,freq=FALSE) # Add 2nd histogram using different color
legend("topright", c("All Days", "Days with WindGustSpeed NA"), fill=c(c1,c2))

#Check if the value of WindGustDir is only unknown on days when wind gust speed is missing.
all(which(df$WindGustDir == "unkn") == which(is.na(df$WindGustSpeed)))
```

WindGustSpeed appears to be missing randomly, other than November 9, 2011 which is also missing some
of the 3pm measurements. WindGustSpeed is only missing on the days when WindGustDir is also unknown.

```{r}
#CHECK NAs FOR CLOUD 3PM
#create variable to identify rows that are missing Cloud3pm
na_Cloud3 <- which(is.na(df_NAs$Cloud3pm))
#check rows of observations missing Cloud3pm
df_NAs[na_Cloud3,]

#CHECK NAs FOR CLOUD 9AM
#create variable to identify rows that are missing Cloud9am
na_Cloud9 <- which(is.na(df_NAs$Cloud9am))
#check rows of observations missing Cloud9am
df_NAs[na_Cloud9,]
```
There are only 3 observations missing of Cloud3pm, and all other variables are complete on those days.
Two of the days are close to each other, both in April 2009.

There is only one observation missing for Cloud9am and it appears to be random. No other values
are missing for the same date.

```{r}
#CHECK NAs FOR MAX & MIN TEMPERATURE

#create variable to identify rows that are missing MaxTemp
na_MaxTemp <- which(is.na(df_NAs$MaxTemp))
#check rows of observations missing MaxTemp
df_NAs[na_MaxTemp,]

#create variable to identify rows that are missing MinTemp
na_MinTemp <- which(is.na(df_NAs$MinTemp))
#check rows of observations missing MinTemp
df_NAs[na_MinTemp,]
```
There is only one observation missing for MaxTemp and it appears to be random. No other values
are missing for the same date. MinTemp is not missing any values in the data set.

```{r}

#CHECK NAs for Temp3pm
#create variable to identify rows that are missing Temp3pm
na_Temp3 <- which(is.na(df_NAs$Temp3pm))
#check rows of observations missing Temp3pm
df_NAs[na_Temp3,]
```
Temperature at 3pm is only missing on November 9, 2011 which is also missing other 3pm measurements, but a
Maximum temperature is available for that day.

```{r}
#CHECK NAs for PRESSURE

#create variable to identify rows that are missing Pressure9am
na_Pres9 <- which(is.na(df_NAs$Pressure9am))
#check rows of observations missing Pressure9am
df_NAs[na_Pres9,]

#create variable to identify rows that are missing Pressure3pm
na_Pres3 <- which(is.na(df_NAs$Pressure3pm))
#check rows of observations missing Pressure9am
df_NAs[na_Pres3,]
```

There is only one observation missing for Pressure9am and it appears to be random. No other values
are missing for the same date.

Pressure3pm is only missing one observation on November 9, 2011 with the other 3pm measurements.

NEXT STEP: IMPUTE MISSING VALUES FROM AIRPORT OR FROM MICE.









