---
title: "Train Test Split"
author: "Kat Weissman"
date: "11/6/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(dplyr)
```


```{r}
#LOAD DATA
dfP <- read.csv("Completed_Location_CSVs/df_Perth_completed.csv", stringsAsFactors = T)
dfC <- read.csv("Completed_Location_CSVs/df_Cairns_completed.csv", stringsAsFactors = T)
dfB <- read.csv("Completed_Location_CSVs/df_Brisbane_completed.csv", stringsAsFactors = T)
#CHANGE VARIABLE TYPE
dfP$Date <- as.Date(dfP$Date)
dfC$Date <- as.Date(dfC$Date)
dfB$Date <- as.Date(dfB$Date)

#sort by date
dfP <- dfP %>% arrange(Date)
dfC <- dfC %>% arrange(Date)
dfB <- dfB %>% arrange(Date)

#create AccuTomorrow factor
accuRainP <- cut(dfP$Rainfall, breaks = c(-1,0.0001,1,15,500), labels = c('NoRain','Mist','Rain','HeavyRain'), include.lowest = TRUE, right = FALSE)
accuRainC <- cut(dfC$Rainfall, breaks = c(-1,0.0001,1,15,500), labels = c('NoRain','Mist','Rain','HeavyRain'))
accuRainB <- cut(dfB$Rainfall, breaks = c(-1,0.0001,1,15,500), labels = c('NoRain','Mist','Rain','HeavyRain'))

table(accuRainP); table(accuRainC); table(accuRainB)

dfP$accuRain <- factor(c(accuRainP[2:length(accuRainP)], factor(ifelse( dfP[ nrow(dfP) ,'RainTomorrow' ] == 'No', 'NoRain', 'Rain'), levels = c('NoRain') ) ), levels = c('1','2','3','4'), labels = c('NoRain','Mist','Rain','HeavyRain') )
dfC$accuRain <- factor(c(accuRainC[2:length(accuRainC)], factor(ifelse( dfC[ nrow(dfC) ,'RainTomorrow' ] == 'No', 'NoRain', 'Rain'), levels = c('NoRain') ) ), levels = c('1','2','3','4'), labels = c('NoRain','Mist','Rain','HeavyRain') )
dfB$accuRain <- factor(c(accuRainB[2:length(accuRainB)], factor(ifelse( dfB[ nrow(dfB) ,'RainTomorrow' ] == 'No', 'NoRain', 'Rain'), levels = c('NoRain') ) ), levels = c('1','2','3','4'), labels = c('NoRain','Mist','Rain','HeavyRain') )

```

For validation of the model, it is necessary to divide the time series sequentially.
The last year of data can be used for validation of the model. June 1, 2013 is the last
date for all locations, therefore a logical division would be from the first date
until May 31, 2012 for the training data sets. The time series for each location 
potentially begins on a different date.

The test data sets will all be the same length from June 1, 2012 until June 1, 2013. 
It is important to test the models on a full year of data due to weather differences 
through the months and seasons.


```{r}
#PERTH DATES
min(dfP$Date)
max(dfP$Date)

#SUBSET DATA BY MAY 31, 2012
train_dfP <- subset(dfP, Date <= "2012-05-31")
test_dfP <- subset(dfP, Date > "2012-05-31")

#PERCENTAGE OF SPLIT
nrow(train_dfP)/nrow(dfP)*100
nrow(test_dfP)/nrow(dfP)*100
```



```{r}
#Brisbane dates
min(dfB$Date)
max(dfB$Date)

#SUBSET DATA BY MAY 31, 2012
train_dfB <- subset(dfB, Date <= "2012-05-31")
test_dfB <- subset(dfB, Date > "2012-05-31")

#PERCENTAGE OF SPLIT
nrow(train_dfB)/nrow(dfB)*100
nrow(test_dfB)/nrow(dfB)*100
```

```{r}
#Cairns dates
min(dfC$Date)
max(dfC$Date)

#SUBSET DATA BY MAY 31, 2012
train_dfC <- subset(dfC, Date <= "2012-05-31")
test_dfC <- subset(dfC, Date > "2012-05-31")

#PERCENTAGE OF SPLIT
nrow(train_dfC)/nrow(dfC)*100
nrow(test_dfC)/nrow(dfC)*100
```

```{r}
summary(train_dfP)
summary(test_dfP)

prop.table(table(train_dfP$RainToday))
prop.table(table(test_dfP$RainToday))
```




```{r}
#write dfs to csv

write.csv(train_dfC,'Train_Test_CSVs/df_Cairns_train.csv', row.names = FALSE)
write.csv(test_dfC,'Train_Test_CSVs/df_Cairns_test.csv', row.names = FALSE)

write.csv(train_dfB,'Train_Test_CSVs/df_Brisbane_train.csv', row.names = FALSE)
write.csv(test_dfB,'Train_Test_CSVs/df_Brisbane_test.csv', row.names = FALSE)

write.csv(train_dfP,'Train_Test_CSVs/df_Perth_train.csv', row.names = FALSE)
write.csv(test_dfP,'Train_Test_CSVs/df_Perth_test.csv', row.names = FALSE)


```



