---
title: 'Group 1: Multivariate analysis of australian climate data'
header-includes:
 - \usepackage{multirow}
 - \usepackage{longtable}
output: 
  pdf_document: 
    latex_engine: xelatex
    includes:
      in_header: 
      - !expr system.file("includes/fig-valign.tex", package = "summarytools")
---


```{r setup, message=FALSE, warning=FALSE, results=FALSE,include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# CLEARING ENVIRONMENT #
rm(list=ls(all=TRUE))


#INSTALL & LOAD LIBRARIES
# Required packages
pkgs<-c("cluster","colorspace","dplyr","factoextra","fpc","ggplot2","ggpubr",
        "NbClust","patchwork","reshape2","stats","summarytools","tidyr","tidyverse",
        "rstudioapi","Clusters")

# Non-installed packages
inspkgs<-pkgs[!pkgs %in% installed.packages()]
for(libs in inspkgs) install.packages(libs, repos = "http://cran.us.r-project.org")

# Loading required
sapply(pkgs,require,character=TRUE)
```

## Data input

To perform the clustering analysis are used the original datasets (numeric variables) for Brisbane, Perth and Cairns.

```{r pressure, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Loading dataset
load('BrisbaneMFA.RData')
dfBrisbane<-df
dfMFABrisbane<-df_MFA
#EigValMFABrisbane<-eig.val
#groupMFABrisbane<-group
#quali.varMFABrisbane<-quali.var
#quanti.varMFABrisbane<-quanti.var
resMFABrisbane<-get_mfa_ind(res.mfa)
coordMFABrisbane<-data.frame(resMFABrisbane$coord)

load('CairnsMFA.RData')
dfCairns<-df
dfMFACairns<-df_MFA
resMFACairns<-get_mfa_ind(res.mfa)
coordMFACairns<-data.frame(resMFACairns$coord)

load('PerthMFA.RData')
dfPerth<-df
dfMFAPerth<-df_MFA
resMFAPerth<-get_mfa_ind(res.mfa)
coordMFAPerth<-data.frame(resMFAPerth$coord)

originaldata<-list(dfMFABrisbane,dfMFAPerth,dfMFACairns)
funum <- function(x){nums<-unlist(lapply(x, is.numeric))
                     x[ , nums]}
orNUMdata<-lapply(originaldata,funum)
names(orNUMdata)<-c("Brisbane","Perth","Cairns")

MFA<-list(coordMFABrisbane,coordMFAPerth,coordMFACairns)
str(MFA)
names(MFA)<-c("MFACoordsBrisbane","MFACoordsPerth","MFACoordsCairns")

listall<-list(orNUMdata$Brisbane,MFA$MFACoordsBrisbane,orNUMdata$Perth,MFA$MFACoordsPerth,orNUMdata$Cairns,MFA$MFACoordsCairns)
names(listall)<-c("Brisbane","MFACoordsBrisbane","Perth","MFACoordsPerth","Cairns","MFACoordsCairns")
str(listall)

set.seed(2021)
```

###  Clustering Analysis

In order to analyze if data presents patterns of association are it is performed a clustering analysis. For this purpose, all incomplete cases remaining are removed and as a first step, the optimal number of clusters are estimated through direct methods: elbow, average silhouette and ASM to choose the most common value of optimal clusters.


```{r, message=FALSE, warnings=FALSE, fig.show="hold", out.width="50%", results=FALSE}
par(mar = c(4,4,.1,.1))
fun01<-function(x){ tmp_df = listall[[x]]
                    tmp_name = names(listall)[x]
                    fviz_nbclust(tmp_df, kmeans, method = "wss") + 
                    geom_vline(xintercept = 4, linetype = 2) + 
                    labs(subtitle = paste("Elbow method",tmp_name))}
fun02<-function(x){ tmp_df = listall[[x]]
                    tmp_name = names(listall)[x]
                    fviz_nbclust(tmp_df, kmeans, method = "silhouette") + 
                    geom_vline(xintercept = 4, linetype = 2) + 
                    labs(subtitle = paste("Elbow method",tmp_name))}
fun03<-function(x){ tmp_df = listall[[x]]
                    tmp_name = names(listall)[x]
                    fviz_nbclust(tmp_df, kmeans, method = "gap_stat") +
                    geom_vline(xintercept = 4, linetype = 2) + 
                    labs(subtitle = paste("Elbow method",tmp_name))}

wss<-lapply(1:length(listall),fun01)
wss

silhouette<-lapply(1:length(listall),fun02)
silhouette

#Gaps<-lapply(1:length(listall),fun03)
#Gaps
```
Given the results provided by the methods, it can be concluded the clustering can be performed with 4 cluster for all the dataset, the original numerical variables and the coordinates of the performed MCA.

```{r, message=FALSE, warnings=FALSE, fig.show="hold", out.width="50%", results=FALSE}
funVizKm<- function(i){ tmp_df = listall[[i]];
                        tmp_kmeans = kmeans(x = listall[[i]], centers = 4)
                        tmp_name = names(listall)[i]
                        fviz_cluster(object = tmp_kmeans, data = listall[[i]], show.clust.cent = TRUE,
                                                   ellipse.type = "euclid", star.plot = TRUE, repel = TRUE) + 
                     theme_bw() + theme(legend.position = "none") + labs(title = 
                                                                                  paste("Results clustering K-means - 4 clusters - ",
                                                                                        tmp_name))}  

VizKmeans<-lapply(1:length(listall),funVizKm) 
VizKmeans

```
```{r, message=FALSE, warnings=FALSE, fig.show="hold", out.width="50%"}
funKm<- function(i){ tmp_df = listall[[i]];
                     tmp_kmeans = kmeans(x = listall[[i]], centers = 4)
                     listall[[i]]<-add_column(listall[[i]], KmeansCluster = tmp_kmeans$cluster)}


Kmeans<-lapply(1:length(listall),funKm)
names(Kmeans)<-c("Brisbane","MFACoordsBrisbane","Perth","MFACoordsPerth","Cairns","MFACoordsCairns")
Kmeans
```

```{r, message=FALSE, warnings=FALSE, fig.show="hold", out.width="50%", results=FALSE}
fun04<-function(x) print(dim(x))
lapply(Kmeans, fun04)

BrisbaneClusters<-cbind(Kmeans[[1]],Kmeans[[2]])
str(BrisbaneClusters)
names(BrisbaneClusters)[24]<-"KmeansClusterMFA"

PerthClusters<-cbind(Kmeans[[3]],Kmeans[[4]])
str(PerthClusters)
names(PerthClusters)[24]<-"KmeansClusterMFA"

CairnsClusters<-cbind(Kmeans[[5]],Kmeans[[6]])
str(CairnsClusters)
names(CairnsClusters)[24]<-"KmeansClusterMFA"

DFClusters<-list(BrisbaneClusters,PerthClusters,CairnsClusters)

```


```{r, message=FALSE, warnings=FALSE, fig.show="hold", out.width="50%", results=FALSE}

fun05<-function(x){ x$KmeansClusterMFA<-as.factor(x$KmeansClusterMFA)
                    levels(x$KmeansClusterMFA)<-list(C1=1,C2=2,C3=3,C4=4)
                    x$KmeansCluster<-as.factor(x$KmeansCluster)
                    levels(x$KmeansCluster)<-list(G1=1,G2=2,G3=3,G4=4)
                    return(x[,-which(names(x) %in% c("Dim.1","Dim.2","Dim.3","Dim.4","Dim.5"))])}

temp<-lapply(DFClusters, fun05)

funConTable<-function(x){ table(x$KmeansClusterMFA,x$KmeansCluster)}
lapply(temp,funConTable)

```

```{r, message=FALSE, warnings=FALSE, fig.show="hold", out.width="50%", results=FALSE}
funProfile<-function(x){catdes(x, num.var=18, prob = 0.01)
                        catdes(x, num.var=19, prob = 0.01)}


lapply(temp,funProfile)
```